---
title: 'Hypothesis Testing: Exercises'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-10-17"
output: html_document
---

# Exercises

## **Exercise 1: One-Sample Z-Test**
1. A manufacturer claims that the average weight of a certain product is 500 grams. A sample of 60 products has an average weight of 510 grams with a population standard deviation of 20 grams. Test the manufacturer's claim at the 5% significance level.
    - State the hypotheses.
    - Perform the one-sample Z-test and make a decision.

```{r}
x_bar <- 510  # Sample mean
mu_0 <- 500   # Population mean under null hypothesis
sigma <- 20   # Population standard deviation
n <- 60       # Sample size
alpha <- 0.05  # Significance level

# H0: 
# H1: 

# TODO Your code here
# Z-test
# Critical value and p-value
```

## **Exercise 2: One-Tailed Z-Test (Right-Tailed)**
2. A researcher believes that the average score on a standardized test is higher than 75. A sample of 40 students has an average score of 78, with a known population standard deviation of 8. Test the researcher’s hypothesis at a 1% significance level (right-tailed test).

```{r}
x_bar <- 78   # Sample mean
mu_0 <- 75    # Population mean under null hypothesis
sigma <- 8    # Population standard deviation
n <- 40       # Sample size
alpha <- 0.01  # Significance level

# H0:
# H1:

# TODO Your code here
```

## **Exercise 3: One-Tailed Z-Test (Left-Tailed)**
3. A company claims that the average lifetime of its light bulbs is at least 1200 hours. A sample of 45 light bulbs has an average lifetime of 1180 hours with a population standard deviation of 50 hours. Test the company's claim at a 10% significance level (left-tailed test).

```{r}
x_bar <- 1180  # Sample mean
mu_0 <- 1200   # Population mean under null hypothesis
sigma <- 50    # Population standard deviation
n <- 45        # Sample size
alpha <- 0.10  # Significance level

# H0: 
# H1: 

# TODO Your code here
```

## **Exercise 4: Two-Sample Z-Test**
4. You are comparing the average heights of male and female students. A sample of 40 male students has an average height of 175 cm, with a standard deviation of 10 cm, and a sample of 35 female students has an average height of 165 cm, with a standard deviation of 8 cm.  
     Test whether there is a significant difference between the average heights of male and female students at a 5% significance level.

```{r}
x_bar_male <- 175   # Sample mean of male students
sd_male <- 10       # Standard deviation of male students
n_male <- 40        # Sample size of male students

x_bar_female <- 165  # Sample mean of female students
sd_female <- 8       # Standard deviation of female students
n_female <- 35       # Sample size of female students

alpha <- 0.05  # Significance level

# H0: 
# H1: 

# TODO Your code here

# Z-test for two independent samples
# Critical value and p-value
```

#### **Exercise 5: Interpretation of p-value**
5. In a hypothesis test for the mean, you obtain a Z-statistic of 1.85. The p-value is calculated to be 0.064.
    - Based on a 5% significance level, would you reject or fail to reject the null hypothesis?
    - What would be your conclusion at a 10% significance level?

```{r}
z <- 1.85
p_value <- 0.064
alpha_5 <- 0.05
alpha_10 <- 0.10

# TODO Your interpretation here
# Decision at 5% significance level
# Decision at 10% significance level
```
