---
title: "R programming basics"
author: "Katerina Vecerkova"
date: "7/11/2024"
output:
  html_document:
    toc: true
---

# Operators

Operators are special symbols or keywords used to perform operations on variables and values. R provides various types of operators to facilitate different kinds of computations and manipulations in your code.

*Types of Operators:*

1. Arithmetic Operators
2. Logical Operators
3. Relational Operators
4. Assignment Operators
5. Miscellaneous Operators

## Arithmetic operators

```{r}
3 + 5  # Addition
8 - 5 # Subtraction
2 * 5  # Multiplication
6 / 2 # Division
2 ^ 3 # Exponentiation
13 %% 4  # Modulus (remainder of the division)
2 ^ 1/2 # Be aware of operator priority
2 ^ (1/2) # And use parentheses
```

## Logical operators

```{r}
TRUE & FALSE  # Logical AND
TRUE | FALSE  # Logical OR
!FALSE  # Logical NOT
(TRUE & TRUE) | (FALSE & FALSE)
xor(TRUE, TRUE) # Exclusive OR
```

## Relational operators

```{r}
FALSE == 0  # Equal to
TRUE == 1  # Equal to
6 != 8  # Not equal to
7 < 8  # Less than
7 > 8  # Greater than
2 <= 7  # Less than or equal to
2 >= 7  # Greater than or equal to
```

## Assignment operators

```{r}
x <- 5  # Left assignment
5 -> x  # Right assignment
x = 5  # Equals assignment (less common)
```

## Miscellaneous operators

```{r}
seq <- 1:5  # Sequence
seq
```

```{r}
df <- data.frame(a = 1:3, b = 4:6)
df$a  # List subset operator
```

```{r}
vec <- c(1, 2, 3)
vec[2]  # Indexing operator
```

```{r}
1 %in% vec  # Membership operator
```

## Special operators

```{r, eval=FALSE}
model <- lm(y ~ x, data = df)  # Tilde operator in a linear model
```

# Variables

Variables are used to store and reuse values in a program. They act as containers for data that can be accessed and manipulated throughout your code. In R, you can assign values to variables using the assignment operator <- (preferred) or (=).

```{r}
myVariable <- "hello" # Camel case naming convention
my_variable <- 3      # Snake case naming convention
my_variable <- 5      # Re-assigning a new value to an existing variable
also_my_variable <- 6 # Another variable following snake case
result <- my_variable + also_my_variable # Using variables in expressions
print(result) # This will output 11
```

# Data types

In R, data types define the kind of data that can be stored and manipulated within a variable. Understanding data types is essential for effective programming and data analysis.

*Basic Data Types:*
1. Numeric: Represents numbers, which can be integers or doubles (real numbers)
2. Integer: Represents whole numbers
3. Character: Represents text or string data
4. Logical: Represents boolean values (`TRUE` or `FALSE`)
5. Complex: Represents complex numbers
6. Factor: Represents categorical data

```{r}
num <- 3.14          # Numeric (double)
int <- 5L            # Integer (note the 'L' suffix)
char <- "hello"      # Character
bool <- TRUE         # Logical
comp <- 1 + 2i       # Complex
fact <- factor("male", levels = c("male", "female"))  # Factor
```

# Data structures

In addition to basic data types, R supports several data structures that can hold multiple values:

1. Vector: A sequence of elements of the same type
2. Matrix: A two-dimensional array of elements of the same type
3. List: A collection of elements, which can be of different types
4. Data Frame: A table-like structure where each column can be of a different type
5. Array: Multi-dimensional generalization of matrices

## Functions to determine data types

1. `class()`: Returns the class/type of an R object.Provides the high-level classification of the object, often corresponding to how the object behaves (e.g., "integer", "numeric", "data.frame").
2. `typeof()`: Returns the internal storage mode of an R object. Gives the low-level internal storage mode of the object (e.g., "integer", "double", "character").
3. `mode()`: Returns the mode of an R object.Similar to `typeof()`, but sometimes gives a higher-level view and is more consistent with older versions of R and S.

```{r}
# Example variable
x <- 42L  # Integer

# Determine class
class_x <- class(x)
print(class_x)

# Determine type
typeof_x <- typeof(x)
print(typeof_x)

# Determine mode
mode_x <- mode(x)
print(mode_x)
```

```{r}
# Example with different data types
y <- "R programming"  # Character

# Determine class
class_y <- class(y)
print(class_y)

# Determine type
typeof_y <- typeof(y)
print(typeof_y)

# Determine mode
mode_y <- mode(y)
print(mode_y)
```

# Functons

Functions are reusable blocks of code that perform specific tasks. They help in breaking down complex problems into smaller, manageable pieces and promote code reusability.

*Why Use Functions?*

- Code Reusability: Functions allow you to write a piece of code once and reuse it multiple times.
- Modularity: Functions help in organizing code into smaller, logical sections.
- Readability: Code becomes easier to read and understand when divided into functions.
- Maintenance: Functions make it easier to update and maintain code.

*Defining a function*

In R, you can define a function using the `function` keyword. A function can have arguments (inputs) and return a value.

```{r, eval=FALSE}
function_name <- function(arg1, arg2, ...) {
  # Function body
  # Perform some operations
  return(value)  # Return a value (optional)
}
```

Let's create a simple function that adds two numbers:

```{r}
add_numbers <- function(a, b) {
  res <- a + b
  return(res)
}

result <- add_numbers(3, 5)
result
```

You can also define functions with default argument values:

```{r}
greet <- function(name = "World") {
  message <- paste("Hello,", name)
  return(message)
}
```

```{r}
greet() # Uses default value
```

```{r}
greet("students!") # Uses provided value
```

Variables defined inside a function are local to that function and are not accessible outside. This is known as the scope of variables.

```{r, eval=FALSE}
increment <- function(x, by = 1) {
  y <- x + by
  return(y)
}

increment(5, 2) # Variable y only exists in the function increment and accessing it will cause an error
```

R also provides a wide range of built-in functions for various tasks like mathematical operations, statistical analysis, and data manipulation. We install packages exactly for the already implemented functions!

```{r}
sqrt(16) # Square root
log(10) # Natural logarithm
mean(c(1, 2, 3, 4, 5)) # Mean of a vector
sd(c(1, 2, 3, 4, 5)) # Standard deviation of a vector
```

# Control structures

## Branching using if and else

The `if` statement allows you to execute code only if a certain condition is true. The `else` statement can be used to execute code if the condition is false.

```{r}
x <- 10

if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is not greater than 5")
}
```

The `ifelse` function is a vectorized version of the `if` statement. It takes three arguments: a condition, a value if the condition is true, and a value if the condition is false.

## Loops

### For

The `for` loop is used to iterate over a sequence of elements.

```{r}
for (i in 1:5) {
  print(i)
}
```

### While

The `while` loop continues to execute as long as a specified condition is true.

```{r}
count <- 1

while (count <= 5) {
  print(count)
  count <- count + 1
}
```

### Repeat

The `repeat` loop executes an infinite loop unless it is interrupted by a `break` statement.

```{r}
count <- 1

repeat {
  print(count)
  count <- count + 1
  if (count > 5) {
    break
  }
}
```

### Switch Statement

The `switch` statement allows you to execute code based on the value of an expression.

```{r}
value <- "b"

result <- switch(value,
                 "a" = "First",
                 "b" = "Second",
                 "c" = "Third",
                 "Unknown")

print(result)
```

# String operations

## Concatenation

The `paste` function concatenates strings with a separator, while `paste0` concatenates strings without a separator.

```{r}
str1 <- "Hello"
str2 <- "World"

# using paste
result_paste <- paste(str1, str2, sep = " ")
print(result_paste)

# using paste0
result_paste0 <- paste0(str1, str2)
print(result_paste0)
```

## Subsetting

The `substr` and `substring` functions extract or replace substrings in a character vector.

```{r}
text <- "Hello World"

# using substr
sub_text <- substr(text, 1, 5)
print(sub_text)

# using substring
sub_text2 <- substring(text, 7)
print(sub_text2)
```

## Length

The `nchar` function returns the number of characters in a string.

```{r}
length_text <- nchar(text)
print(length_text)
```

## Replacement

The `sub` function replaces the first instance of a pattern, while `gsub` replaces all instances.

```{r}
text <- "The rain in Spain stays mainly in the plain"

# using sub
result_sub <- sub("ain", "xxx", text)
print(result_sub)

# using gsub
result_gsub <- gsub("ain", "xxx", text)
print(result_gsub)
```

## Pattern Matching

The `grep` function searches for patterns and returns the indices, while `grepl` returns a logical vector indicating if a pattern was found.

```{r}
fruit <- c("apple", "banana", "cherry")

# using grep
result_grep <- grep("a", fruit)
print(result_grep)

# using grepl
result_grepl <- grepl("a", fruit)
print(result_grepl)
```

Functions `regexpr`, `gregexpr` and `regexc` provide more detailed pattern matching information.

```{r, eval=FALSE}
?regex
```

## Splitting

The `strsplit` function splits strings into substrings based on a delimiter.

```{r}
result_strsplit <- strsplit(text, " ")
print(result_strsplit)
```

## Case Conversion

The `toupper` and `tolower` functions convert strings to uppercase and lowercase, respectively.

```{r}
# using toupper
upper_text <- toupper(text)
print(upper_text)

# using tolower
lower_text <- tolower(text)
print(lower_text)
```

## Trimming

The `trimws` function removes leading and trailing whitespace from strings.

```{r}
text <- "   Hello World   "

trimmed_text <- trimws(text)
print(trimmed_text)
```
