---
title: 'R programming basic: Exercises Solutions'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exercises

## Arithmetic Operators

1. **Addition Exercise**
   - Calculate the result of adding 12 and 8.

```{r}
12 + 8
```

2. **Modulus Exercise**
   - Find the remainder when 29 is divided by 6.

```{r}
29 %% 6
```

## Logical Operators

1. **Logical AND Exercise**
   - Evaluate the result of `TRUE & FALSE`.

```{r}
TRUE & FALSE
```

2. **Logical OR Exercise**
   - Determine the result of `TRUE | FALSE`.

```{r}
TRUE | FALSE
```

## Relational Operators

1. **Equality Check**
   - Check if `7` is equal to `5`.
   
```{r}
7 == 5
```

2. **Greater Than Check**
   - Verify if `10` is greater than `3`.

```{r}
10 > 3
```

## Assignment Operators

1. **Left Assignment**
   - Assign the value `15` to a variable named `x` using the left assignment operator `<-`.

```{r}
x <- 15
```

2. **Right Assignment**
   - Assign the value `20` to a variable named `y` using the right assignment operator `->`.

```{r}
20 -> y
```

## Miscellaneous Operators

1. **Sequence Creation**
   - Create a sequence from 1 to 10 using the `:` operator.
   
```{r}
1:10
```

2. **List Subsetting**
   - Access the first column of a data frame `df` where `df` is created with `a = 1:3` and `b = 4:6`.

```{r}
df <- data.frame(a = 1:3, b = 4:6)
df$a
```

## Variables

1. **Variable Assignment**
   - Assign the value `7` to a variable named `result` and then print it.

```{r}
result <- 7
print(result)
```

2. **Variable Re-assignment**
   - Reassign the value `10` to the variable `result` and print the updated value.

```{r}
result <- 10
print(result)
```

## Data Types

1. **Identify Data Type**
   - Check the type of the variable `num` which is assigned the value `3.14`.
   - Hint: Use `typeof()`.

```{r}
num <- 3.14
typeof(num)
```

2. **Create a Factor**
   - Create a factor variable `fact` with levels `male` and `female`, and assign it the value `female`.

```{r}
factor("female", levels = c("female", "male"))
```

## Functions

1. **Create a Function**
   - Define a function `multiply` that takes two arguments and returns their product.
   - Hint: Use `function()` to define the function.

```{r}
multiply <- function(a, b) {
  return(a * b)
}
```

2. **Function with Default Argument**
   - Define a function `subtract` with default argument value `10`. It should subtract a given number from this default value.
   - Hint: Provide a default value for the argument in the function definition.

```{r}
subtract <- function(x = 10) {
  return(10 - x)
}
```

## Control Structures

1. **If-Else Statement**
   - Write an `if` statement to check if `x` is less than `10`, and print "x is less than 10" if true, otherwise print "x is 10 or more".
   - Hint: Use `if` and `else`.

```{r}
x <- 8
if (x < 10) {
  print("x is less than 10")
} else {
  print("x is 10 or more")
}
```

2. **For Loop**
   - Write a `for` loop to print numbers from `1` to `3`.

```{r}
for (i in 1:3) {
  print(i)
}
```

## String Operations

1. **String Concatenation**
   - Concatenate the strings `"Data"` and `"Science"` with a space in between.

```{r}
str1 <- "Data"
str2 <- "Science"
paste(str1, str2)
```

2. **Substring Extraction**
   - Extract the substring from position `1` to `4` of the string `"Programming"`.

```{r}
text <- "Programming"
substr(text, 1, 4)
```

3. **String Replacement**
   - Replace all occurrences of `"the"` with `"a"` in the string `"The quick brown fox jumps over the lazy dog"`.

```{r}
text <- "The quick brown fox jumps over the lazy dog"
gsub("the", "a", text, ignore.case = TRUE)
```

4. **Case Conversion**
   - Convert the string `"hello world"` to uppercase.

```{r}
text <- "hello world"
toupper(text)
```

5. **String Splitting**
   - Split the string `"apple,banana,cherry"` by commas.

```{r}
text <- "apple,banana,cherry"
strsplit(text, ",")
```