---
title: 'Intro to R: Exercises'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exercises

## Exercise 1: Installing and Loading Packages

### Task:
1. **Install the `ggplot2` and `dplyr` packages**
2. **Load these packages**

```{r, eval=FALSE}
# install packages
install.packages("ggplot2")
install.packages("dplyr")

# load the installed packages so we can use the implemented functions
library(ggplot)
library(dplyr)
```

## Exercise 2: Navigating RStudio

### Task:
1. **Run the following command in RStudio** to access the documentation for the `nrow` function:

```{r}
?nrow
```

2. **Create a new R script** in RStudio and run the entire script using the Run All option.

## Exercise 3: Working with R Projects and `renv`

### Task:
1. **Initialize a new project with a `renv` environment**

   **Hint:** Use `renv::init()` to initialize a new environment or check the box in the dialog window when creating new R project.

2. **Install the `ggplot2` and `dplyr` packages** within this `renv` environment.

3. **Take a snapshot of the current environment** to create an `renv.lock` file:

```{r, eval=FALSE}
install.packages("ggplot2")
install.packages("dplyr")

renv::snapshot()
```

4. **Restore the environment** from the `renv.lock` file to ensure consistency.

```{r}
renv::restore()
```

## Exercise 4: Creating an R Markdown Document

### Task:
1. **Write a simple R Markdown document** that includes:
   - A title and author in the YAML header.
   - A section with a heading and a paragraph describing what R is.
   - An R code chunk that calculates and displays the mean of a numeric vector.

   **Hint:** Use `rmarkdown::render()` to generate the document from your R Markdown file.

   ```markdown
   ---
   title: "Simple Report"
   author: "Your Name"
   output: html_document
   ---

   ## Introduction to R

   R is a programming language used for statistical analysis.

   ```{r}
   # Numeric vector
   numbers <- c(1, 2, 3, 4, 5)
   
   # Calculate mean
   mean_value <- mean(numbers)
   mean_value
   ```
   ```

## (Optional) Exercise 5: Building a Shiny App

### Task:
1. **Create a basic Shiny app** with a slider input to select a range and a plot output that displays a line plot based on the selected range.

   **Hint:** Define `ui` for the user interface and `server` for server-side logic. Use `shinyApp()` to run the app.

```{r}
# Load the Shiny package
library(shiny)

# Define UI
ui <- fluidPage(
  
  # Application title
  titlePanel("Shiny App with Slider and Plot"),
  
  # Sidebar layout with input and output
  sidebarLayout(
    sidebarPanel(
      # Slider input: Select a range of values
      sliderInput("range", 
                  "Select Range:", 
                  min = 1, 
                  max = 100, 
                  value = c(10, 50))
    ),
    
    # Show the plot
    mainPanel(
      plotOutput("linePlot")
    )
  )
)

# Define Server logic
server <- function(input, output) {
  
  output$linePlot <- renderPlot({
    # Generate x values based on the selected range
    x <- seq(input$range[1], input$range[2], by = 1)
    y <- sin(x)  # Example: Using the sine function for plotting
    
    # Create the plot
    plot(x, y, type = "l", col = "blue", lwd = 2,
         xlab = "X values", ylab = "Y values (sin(x))",
         main = "Line Plot of sin(x)")
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```