---
title: 'Intro to R: Exercises'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "9/12/2024"
output: html_document
---

# Exercises

## Exercise 1: Installing and Loading Packages

### Task:
1. **Install the `ggplot2` and `dplyr` packages**
2. **Load these packages**

```{r}
# TODO your code here
```

## Exercise 2: Navigating RStudio

### Task:
1. **Run the following command in RStudio** to access the documentation for the `nrow` function:

```{r}
# TODO your code here
```

2. **Create a new R script** in RStudio and run the entire script using the Run All option.

## Exercise 3: Working with R Projects and `renv`

### Task:
1. **Initialize a new project with a `renv` environment**

   **Hint:** Use `renv::init()` to initialize a new environment or check the box in the dialog window when creating new R project.

2. **Install the `ggplot2` and `dplyr` packages** within this `renv` environment.

3. **Take a snapshot of the current environment** to create an `renv.lock` file:

```{r}
# TODO your code here
```

4. **Restore the environment** from the `renv.lock` file to ensure consistency.

```{r}
# TODO your code here
```

## Exercise 4: Creating an R Markdown Document

### Task:
1. **Write a simple R Markdown document** that includes:
   - A title and author in the YAML header.
   - A section with a heading and a paragraph describing what R is.
   - An R code chunk that calculates and displays the mean of a numeric vector.

   **Hint:** Use `rmarkdown::render()` to generate the document from your R Markdown file.

   ```markdown
   ---
   title: "Simple Report"
   author: "Your Name"
   output: html_document
   ---

   ## Introduction to R

   R is a programming language used for statistical analysis.

   ```{r}
   # Numeric vector
   numbers <- c(1, 2, 3, 4, 5)
   
   # Calculate mean
   mean_value <- mean(numbers)
   mean_value
   ```
   ```

## Exercise 5: Building a Shiny App

### Task:
1. **Create a basic Shiny app** with a slider input to select a range and a plot output that displays a line plot based on the selected range.

   **Hint:** Define `ui` for the user interface and `server` for server-side logic. Use `shinyApp()` to run the app.

```{r}
# TODO your code here
```