---
title: "Parameter Estimation: Solution"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction to Point Estimation

Point estimation is a statistical method used to estimate an unknown population parameter based on sample data. A point estimator provides a single value as an estimate of a parameter. This value is called a point estimate.

## Key Concepts

1. **Population Parameter**: A value that describes a characteristic of a population, such as the mean or variance. It is usually unknown.
2. **Sample Statistic**: A value calculated from sample data that serves as an estimate of the population parameter.
3. **Point Estimator**: A rule or formula used to calculate the sample statistic that estimates the population parameter.

Common point estimators include:

  - **Sample Mean**: An estimate of the population mean.
  - **Sample Variance**: An estimate of the population variance.

In this tutorial, we will explore how to calculate and interpret these common point estimators using R.

# Examples of Point Estimation

We will use the built-in `mtcars` dataset in R for our examples. This dataset contains information about various car models, including miles per gallon (mpg), number of cylinders, horsepower, and more.

## Setting Up

```{r setup, include=FALSE}
# load necessary libraries
library(ggplot2)
library(dplyr)

# load the mtcars dataset
data(mtcars)
```

## Estimating the Population Mean

The sample mean is a point estimate of the population mean. It is calculated by summing all sample values and dividing by the number of values.

### Example

Calculate the sample mean of the `mpg` (miles per gallon) variable.

```{r}
# calculate sample mean of the 'mpg' variable
mean_mpg <- mean(mtcars$mpg)
mean_mpg
```

### Exercise 1

Calculate the sample mean of the `hp` (horsepower) variable in the `mtcars` dataset.

```{r}
mean(mtcars$hp)
```

## Estimating the Population Variance

The sample variance is a point estimate of the population variance. It measures the dispersion of the sample values from the sample mean.

### Example

Calculate the sample variance of the `mpg` variable.

```{r}
# calculate sample variance of the 'mpg' variable
var_mpg <- var(mtcars$mpg)
var_mpg
```

### Exercise 2

Calculate the sample variance of the `wt` (weight) variable in the `mtcars` dataset.

```{r}
var(mtcars$mpg)
```

### Exercise 3

Calculate the sample mean and sample variance of the `qsec` (1/4 mile time) variable in the `mtcars` dataset.

```{r}
# calculate sample mean and variance of the 'qsec' variable
mean(mtcars$qsec)
var(mtcars$qsec)
```

