---
title: 'Exploratory Data Analysis: Exercises Solutions'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exercises Solutions

## Understanding and Summarizing the Data

### Exercise 1: Load and Preview the Data

**Task:** Load the Titanic dataset from a CSV file and display the first 10 rows to familiarize yourself with the data.

```{r}
library(tidyverse, quietly = TRUE)
data <- read.csv("titanic.csv")
head(data, n=10)
```

### Exercise 2: Data Structure and Summary Statistics

**Task:** 
1. Use the `str()` function to examine the data structure, identifying data types for each column.
2. Obtain a summary of all variables using `summary()`. Identify any interesting values or outliers.

```{r}
str(data)
```
```{r}
summary(data)
```

### Exercise 3: Checking for Missing Values

**Task:** Use the `is.na()` function to check for missing values. Summarize which columns contain missing values and their frequency.

```{r}
colSums(is.na(data))
```

## Visualizing Key Variables

### Exercise 4: Age Distribution

**Task:** Create a histogram of the `Age` column to understand its distribution. Choose an appropriate number of bins (e.g., 20) and label your axes.

```{r}
library(ggplot2, quietly = TRUE)
ggplot(data, aes(x=Age)) + geom_histogram(bins = 20, color = "darkolivegreen", fill = "darkolivegreen1") + xlab("Age") + ylab("Frequency")
```

### Exercise 5: Survival Rates by Gender

**Task:** Create a bar plot showing the survival rates (`Survived`) by gender (`Sex`). This will help in understanding the impact of gender on survival.

```{r}
ggplot(data, aes(x = Sex, y = Survived)) +
  geom_col(color = "orange3", fill = "orange2") +
  labs(title = "Barplot of Gender by Survived", x = "Survived", y = "Gender")
```

### Exercise 6: Fare Distribution by Passenger Class

**Task:** Create a boxplot showing the distribution of `Fare` by passenger class (`Pclass`). Interpret whether fare distributions differ across classes.

```{r}
ggplot(data, aes(x = as.factor(Pclass), y = Fare)) +
  geom_boxplot(fill = "tomato1") +
  labs(title = "Boxplot of Age by Survived", x = "Passenger Class", y = "Fare")
```

## Handling Missing Data

### Exercise 7: Imputing Missing Age Values

**Task:** Fill missing values in the `Age` column with the median age. Confirm that there are no remaining missing values.

```{r}
# from the histogram above we can see that the distribution of Age is skewed to the left, therefore we are going to imput with the column median
age_median <- median(data$Age, na.rm = TRUE)
data <- data %>% mutate(Age = replace_na(Age, age_median))

# check the count is NAs after imputation
sum(is.na(data$Age))
```

### Exercise 8: Dealing with Missing Embarked Data

**Task:** Replace missing values in the `Embarked` column with the most common embarkation port. Display the column to confirm the change.

```{r}
embarked_without_na <- data %>% select(Embarked) %>% drop_na()
frequency_table <- table(embarked_without_na)
embarked_mode <- names(frequency_table)[which.max(frequency_table)]
data <- data %>% mutate(Embarked = replace_na(Embarked, embarked_mode))

# check the count is NAs after imputation
sum(is.na(data$Embarked))
```

## Basic Bivariate Analysis

### Exercise 9: Exploring Fare and Survival Relationship

**Task:** Create a boxplot to examine the relationship between `Fare` and `Survived`. What does this visualization suggest about the effect of fare on survival?

```{r}
ggplot(data, aes(x = as.factor(Survived), y = Fare)) +
  geom_boxplot(fill = "firebrick4") +
  labs(title = "Boxplot of Fare by Survived", x = "Survived", y = "Fare")
```

### Exercise 10: Correlation Analysis

**Task:** Create a correlation matrix for the numerical columns (e.g., `Age`, `Fare`, `SibSp`, and `Parch`). Interpret any strong positive or negative correlations.

```{r}
library(reshape2, quietly = TRUE)
correlation_matrix <- round(cor(data %>% select(Survived, Pclass, Age, SibSp, Parch, Fare)), 2)
correlation_matrix
melted_correlation_matrix <- melt(correlation_matrix)
melted_correlation_matrix
ggplot(data = melted_correlation_matrix, aes(x = Var1, y = Var2, fill = value)) + geom_tile() + xlab("") + ylab("")
```

## Feature Engineering

### Exercise 11: Extracting Title from Name

**Task:** Extract the title (e.g., Mr., Mrs., Miss.) from the `Name` column and create a new `Title` column. Group rare titles (those that appear less than 5 times) into an "Other" category. Display a summary of the new `Title` column.

```{r}
# look at the structure of the Name column
head(data$Name)

data <- data |> 
  separate(Name, into = c("last_name", "title_and_first_name"), sep = ", ", remove = FALSE) |> 
  separate(title_and_first_name, into = c("title", "first_name"), sep = " ", extra = "merge") |> 
  select(-c(last_name, first_name))

# check the extracted titles
head(data$title)

# looks at the frequencies of titles
titles_frequency_table <- table(data$title)
titles_frequency_table

# replace titles with frequency < 5 with "Other"
rare_names <- names(titles_frequency_table[titles_frequency_table < 5])

data <- data |> 
  mutate(title = if_else(title %in% rare_names, "Other", title))

# check the result
table(data$title)
```

### Exercise 12: Creating a Family Size Variable

**Task:** Create a new feature called `FamilySize` by combining `SibSp` (siblings/spouses) and `Parch` (parents/children) to represent the total family size onboard. Display a summary of the new column.

```{r}
data <- data |>
  mutate(FamilySize = SibSp + Parch)

summary(data$FamilySize)
```

## Discussion

### Exercise 13: Observations and Summary

**Task:** Briefly summarize your observations from the EDA. Consider any notable patterns, potential outliers, or areas for further investigation.

```{r}
# TODO your discussion here
```

