---
title: 'Exploratory Data Analysis: Exercises'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exercises

## Understanding and Summarizing the Data

### Exercise 1: Load and Preview the Data

**Task:** Load the Titanic dataset from a CSV file and display the first 10 rows to familiarize yourself with the data.

```{r}
# TODO your code here
```

### Exercise 2: Data Structure and Summary Statistics

**Task:** 
1. Use the `str()` function to examine the data structure, identifying data types for each column.
2. Obtain a summary of all variables using `summary()`. Identify any interesting values or outliers.

```{r}
# TODO your code here
```

### Exercise 3: Checking for Missing Values

**Task:** Use the `is.na()` function to check for missing values. Summarize which columns contain missing values and their frequency.

```{r}
# TODO your code here
```

## Visualizing Key Variables

### Exercise 4: Age Distribution

**Task:** Create a histogram of the `Age` column to understand its distribution. Choose an appropriate number of bins (e.g., 20) and label your axes.

```{r}
# TODO your code here
```

### Exercise 5: Survival Rates by Gender

**Task:** Create a bar plot showing the survival rates (`Survived`) by gender (`Sex`). This will help in understanding the impact of gender on survival.

```{r}
# TODO your code here
```

### Exercise 6: Fare Distribution by Passenger Class

**Task:** Create a boxplot showing the distribution of `Fare` by passenger class (`Pclass`). Interpret whether fare distributions differ across classes.

```{r}
# TODO your code here
```

## Handling Missing Data

### Exercise 7: Imputing Missing Age Values

**Task:** Fill missing values in the `Age` column with the median age. Confirm that there are no remaining missing values.

```{r}
# TODO your code here
```

### Exercise 8: Dealing with Missing Embarked Data

**Task:** Replace missing values in the `Embarked` column with the most common embarkation port. Display the column to confirm the change.

```{r}
# TODO your code here
```

## Basic Bivariate Analysis

### Exercise 9: Exploring Fare and Survival Relationship

**Task:** Create a boxplot to examine the relationship between `Fare` and `Survived`. What does this visualization suggest about the effect of fare on survival?

```{r}
# TODO your code here
```

### Exercise 10: Correlation Analysis

**Task:** Create a correlation matrix for the numerical columns (e.g., `Age`, `Fare`, `SibSp`, and `Parch`). Interpret any strong positive or negative correlations.

```{r}
# TODO your code here
```

## Feature Engineering

### Exercise 11: Extracting Title from Name

**Task:** Extract the title (e.g., Mr., Mrs., Miss.) from the `Name` column and create a new `Title` column. Group rare titles (those that appear less than 5 times) into an "Other" category. Display a summary of the new `Title` column.

```{r}
# TODO your code here
```

### Exercise 12: Creating a Family Size Variable

**Task:** Create a new feature called `FamilySize` by combining `SibSp` (siblings/spouses) and `Parch` (parents/children) to represent the total family size onboard. Display a summary of the new column.

```{r}
# TODO your code here
```

## Discussion

### Exercise 13: Observations and Summary

**Task:** Briefly summarize your observations from the EDA. Consider any notable patterns, potential outliers, or areas for further investigation.

```{r}
# TODO your discussion here
```
