---
title: "Exploratory Data Analysis"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exploratory Data Analysis

## **1. What is Exploratory Data Analysis (EDA)?**
   - **Definition**: EDA is an approach to analyzing data sets to summarize their main characteristics, often using visual methods.
   - **Purpose**: Understand the data structure, detect outliers or anomalies, recognize patterns, and generate hypotheses.
   - **Importance**: Before performing more complex statistical analysis, EDA helps ensure data quality and identify assumptions that may impact later stages of analysis.

#### **2. Steps in EDA**

1. **Understanding Data Types and Structure**
   - **Types of Data**: Categorical (nominal, ordinal) vs. Numeric (continuous, discrete).
   - **Data Structure in R**: Briefly cover data frames, factors, lists, and numerical data types (integers, doubles).

```{r}
# Loading and viewing data structure
data <- read.csv("titanic.csv")
str(data)
summary(data)
```

2. **Summary Statistics**
   - **Numerical Variables**:
      - Mean, Median, Standard Deviation, Range, Minimum, Maximum, Quartiles.
      - Useful functions: `mean()`, `median()`, `sd()`, `range()`, `summary()`.

```{r}
# Summary statistics
summary(data$Fare)
mean(data$Fare, na.rm = TRUE)
```

   - **Categorical Variables**:
      - Frequency counts and proportions.
      - Useful functions: `table()`, `prop.table()`.

```{r}
# Frequency table
table(data$Cabin)
```

3. **Data Visualization Techniques**
   - **Histogram**: Visualize the distribution of a single continuous variable.
```{r}
library(ggplot2)

ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 15, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Variable", x = "Values", y = "Frequency")
```
   
   - **Box Plot**: Identify the distribution, central tendency, and potential outliers for a variable, often grouped by category.
   
```{r}
ggplot(data, aes(x = factor(Survived), y = Age)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Age by Survived", x = "Survived", y = "Age")
```
   
   - **Bar Plot**: Display frequencies of categorical variables.
   
```{r}
ggplot(data, aes(x = Embarked)) +
  geom_bar(fill = "lightcoral") +
  labs(title = "Bar Plot of Embarked", x = "Embarked", y = "Count")
```
   
   - **Scatter Plot**: Show the relationship between two continuous variables.
   
```{r}
ggplot(data, aes(x = Age, y = Fare)) +
  geom_point(color = "steelblue") +
  labs(title = "Scatter Plot of Age vs Fare", x = "Age", y = "Fare")
```

4. **Detecting Outliers**
   - **Outliers**: Data points significantly different from others in the dataset.
   - **Identification Methods**: Boxplots, z-scores (standardization), and IQR method.
```{r}
# Identifying outliers using IQR
Q1 <- quantile(data$Age, 0.25, na.rm=TRUE)
Q3 <- quantile(data$Age, 0.75, na.rm=TRUE)
IQR <- Q3 - Q1
outliers <- data$Age[data$Age < (Q1 - 1.5 * IQR) | data$Age > (Q3 + 1.5 * IQR)]
outliers <- outliers[!is.na(outliers)] # remove NAs
outliers
```

5. **Handling Missing Values**
   - **Assess Missingness**: Check for any missing values in the data and the proportion of missingness.
```{r}
# Check for missing values
colSums(is.na(data))
```
   - **Dealing with Missing Values**: Common approaches include removing or imputing (mean/median) missing values.

6. **Exploring Relationships Between Variables**
   - **Correlation Matrix**: Identify linear relationships between pairs of continuous variables.
```{r}
# Calculate correlation matrix
library(reshape2)
cor_matrix <- cor(data[, sapply(data, is.numeric)], use = "complete.obs")
cor_data <- melt(cor_matrix) # reshape the matrix for ggplot2
ggplot(cor_data, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Heatmap of Correlation Matrix", x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
   
   - **Cross-tabulation**: Analyze the relationship between two categorical variables.
   
```{r}
# Cross-tabulation
table(data$Survived, data$Sex)
```

  - **Principal Component Analysis (PCA)**: is a dimensionality reduction technique that transforms data into a set of orthogonal components (principal components) that capture the most variance in the data. By projecting the data onto these components, PCA reduces the number of variables while preserving as much of the original variability as possible. This is useful for simplifying datasets, revealing underlying structures, and visualizing high-dimensional data in lower dimensions.
  
```{r}
library(magrittr)

# select numeric columns for PCA
numeric_columns <- data %>%
  dplyr::select(Age, Fare, SibSp, Parch, Survived) %>%
  na.omit()  # remove rows with missing values

# perform PCA
pca_result <- prcomp(numeric_columns, scale. = TRUE)
pca_scores <- as.data.frame(pca_result$x)

# add the survived column to the pca scores
pca_scores$survived <- numeric_columns$Survived
```

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7, aes(color = survived)) +
  labs(title = "PC1 vs PC2", x = "PC1", y = "PC2") +
  theme_minimal()
```

```{r}
ggplot(pca_scores, aes(x = PC2, y = PC3)) +
  geom_point(alpha = 0.7, aes(color = survived)) +
  labs(title = "PC2 vs PC3", x = "PC2", y = "PC3") +
  theme_minimal()
```

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC3)) +
  geom_point(alpha = 0.7, aes(color = survived)) +
  labs(title = "PC1 vs PC3", x = "PC1", y = "PC3") +
  theme_minimal()
```

#### **3. Key Insights and Next Steps**
   - **Summarizing Findings**: Briefly describe key insights, unusual patterns, or any issues like skewness or high missingness.
   - **Generating Hypotheses**: Based on EDA, suggest potential hypotheses or questions for further analysis.

