---
title: "Exploratory Data Analysis"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-10-31"
output: html_document
---

# Exploratory Data Analysis

## **1. What is Exploratory Data Analysis (EDA)?**
   - **Definition**: EDA is an approach to analyzing data sets to summarize their main characteristics, often using visual methods.
   - **Purpose**: Understand the data structure, detect outliers or anomalies, recognize patterns, and generate hypotheses.
   - **Importance**: Before performing more complex statistical analysis, EDA helps ensure data quality and identify assumptions that may impact later stages of analysis.

#### **2. Steps in EDA**

1. **Understanding Data Types and Structure**
   - **Types of Data**: Categorical (nominal, ordinal) vs. Numeric (continuous, discrete).
   - **Data Structure in R**: Briefly cover data frames, factors, lists, and numerical data types (integers, doubles).

```{r}
# Example: Loading and viewing data structure
data <- read.csv("titanic.csv")
str(data)
summary(data)
```

2. **Summary Statistics**
   - **Numerical Variables**:
      - Mean, Median, Standard Deviation, Range, Minimum, Maximum, Quartiles.
      - Useful functions: `mean()`, `median()`, `sd()`, `range()`, `summary()`.

```{r}
# Example: Summary statistics
summary(data$Fare)
mean(data$Fare, na.rm = TRUE)
```

   - **Categorical Variables**:
      - Frequency counts and proportions.
      - Useful functions: `table()`, `prop.table()`.

```{r}
# Example: Frequency table
table(data$Cabin)
```

3. **Data Visualization Techniques**
   - **Histogram**: Visualize the distribution of a single continuous variable.
```{r}
library(ggplot2)

ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 15, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Variable", x = "Values", y = "Frequency")
```
   - **Box Plot**: Identify the distribution, central tendency, and potential outliers for a variable, often grouped by category.
```{r}
ggplot(data, aes(x = factor(Survived), y = Age)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Age by Survived", x = "Survived", y = "Age")
```
   - **Bar Plot**: Display frequencies of categorical variables.
```{r}
ggplot(data, aes(x = Embarked)) +
  geom_bar(fill = "lightcoral") +
  labs(title = "Bar Plot of Embarked", x = "Embarked", y = "Count")
```
   - **Scatter Plot**: Show the relationship between two continuous variables.
```{r}
ggplot(data, aes(x = Age, y = Fare)) +
  geom_point(color = "steelblue") +
  labs(title = "Scatter Plot of Age vs Fare", x = "Age", y = "Fare")
```

4. **Detecting Outliers**
   - **Outliers**: Data points significantly different from others in the dataset.
   - **Identification Methods**: Boxplots, z-scores (standardization), and IQR method.
```{r}
# Example: Identifying outliers using IQR
Q1 <- quantile(data$Age, 0.25, na.rm=TRUE)
Q3 <- quantile(data$Age, 0.75, na.rm=TRUE)
IQR <- Q3 - Q1
outliers <- data$Age[data$Age < (Q1 - 1.5 * IQR) | data$Age > (Q3 + 1.5 * IQR)]
outliers <- outliers[!is.na(outliers)] # remove NAs
outliers
```

5. **Handling Missing Values**
   - **Assess Missingness**: Check for any missing values in the data and the proportion of missingness.
```{r}
# Example: Check for missing values
colSums(is.na(data))
```
   - **Dealing with Missing Values**: Common approaches include removing or imputing (mean/median) missing values.

6. **Exploring Relationships Between Variables**
   - **Correlation Matrix**: Identify linear relationships between pairs of continuous variables.
```{r}
# Example: Calculate correlation matrix
library(reshape2)
cor_matrix <- cor(data[, sapply(data, is.numeric)], use = "complete.obs")
cor_data <- melt(cor_matrix) # reshape the matrix for ggplot2
ggplot(cor_data, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Heatmap of Correlation Matrix", x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
   - **Cross-tabulation**: Analyze the relationship between two categorical variables.
```{r}
# Example: Cross-tabulation
table(data$Survived, data$Sex)
```

#### **3. Key Insights and Next Steps**
   - **Summarizing Findings**: Briefly describe key insights, unusual patterns, or any issues like skewness or high missingness.
   - **Generating Hypotheses**: Based on EDA, suggest potential hypotheses or questions for further analysis.

