---
title: "Tidyverse Data Manipulation"
author: "Kateřina Večerková"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
---

# `tidyverse` data manipulation

The `tidyverse` is a collection of R packages developed for data science. All `tidyverse` packages share underlying ideas, syntax and data structures, making them highly interopable. The `tidyverse` simplifies many common data manipulation, visualization and analysis tasks. `tidyverse` packages include:
- `readr`: data import
- `dplyr`: data manipulation
- `tidyr`: data tidying
- `ggplot2`: data visualization
- `forcats`: categorical data manipulation
- `stringr`: string manipulation
- `lubridate`: date and time manipulation
- `tibble`: modern data frames
- `purrr`: functional programming

To install the complete tidyverse, use
```{r, message=FALSE, warning=FALSE, results='hide'}
install.packages("tidyverse")
library(tidyverse)
```

## Loading data from files with readr

The `readr` R package offers functions for importing and exporting tabular data in various formats, such as comma-separated values (CSV), tab-separated values (TSV), and other delimiter-separated files.

```{bash}
head iris.csv
```

CSV, which stands for comma-separated values, is a widely used format for storing tabular data. The first row typically contains the column names, and each subsequent row contains the corresponding data values separated by commas. We're gonna load this dataset form the file `iris.csv` to R.


```{r, message=FALSE, warning=FALSE, results='hide'}
# import the readr package
library(readr)
# load the csv dataset
iris <- readr::read_csv("iris.csv")
# let's look at the first 5 rows
head(iris)
```

Another widely used format is TSV, or tab-separated values. In this case the delimier of columns isn't a comma, but a tab `\t`.

```{r}
# laod the tsv dataset
iris <- readr::read_tsv("iris.tsv")
head(iris)
```

## Manipulating data with dplyr

```{r, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
```

### 1. Filtering Rows
Filter the dataset to include only rows where the species is "setosa".

```{r}
setosa_iris <- iris %>% dplyr::filter(variety == "setosa")
head(setosa_iris)
```

Notice the %>% (pipe) operator, it expresses a sequence of multiple operations. Also it is not necessary to put the package before the function, here I explicitly wrote it to demonstrate the use of different tidyverse packages.

### 2. Selecting Columns
Select only the columns for sepal length and sepal width.

```{r}
sepal_data <- iris %>% dplyr::select(sepal.length, sepal.width)
head(sepal_data)
```
### 3. Adding New Columns
Create a new column that is the ratio of petal length and petal width.

```{r}
iris <- iris %>% dplyr::mutate(petal.ratio = petal.length / petal.width)
head(iris)
```

### 4. Summarizing data
Calculate the mean and standard deviation of sepal length for each species.

```{r}
summary_stats <- iris %>% dplyr::group_by(variety) %>% dplyr::summarize(mean_sepal.length = mean(sepal.length), sd_sepal.length = sd(sepal.length))
summary_stats
```

### 5. Arranging rows
Arrange the dataset by sepal length in descending order.

```{r}
sorted_iris <- iris %>% dplyr::arrange(desc(sepal.length))
head(sorted_iris)
```

### 6. Grouping and Summarizing
For each species, count the number of instances where petal length is greater than 4.

```{r}
count_petal_length <- iris %>% dplyr::filter(petal.length > 4) %>% dplyr::group_by(variety) %>% dplyr::summarize(count = n())
count_petal_length
```
### 7. Renaming Columns
Rename the columns sepal.length and sepal.width to sepal_length and sepal_width.

```{r}
iris <- iris %>% dplyr::rename(sepal_length=sepal.length, sepal_width=sepal.width)
colnames(iris)
```

### 8. Filtering and Mutating
Filter the dataset to include only the "virginica" species and ten add a column indicating whether sepal length is greater than 6.

```{r}
virginica_iris <- iris %>% dplyr::filter(variety == "virginica") %>% dplyr::mutate(sepal_length_over_6 = sepal_length > 6)
head(virginica_iris)
```
### 9. Sampling Rows
Randomly sample 10 rows from the dataset.

```{r}
sampled_iris <- iris %>% dplyr::sample_n(10)
show(sampled_iris)
```
### 10. Combining Multiple dplyr Functions
Create a new dataset that includes only the "versicolor" species, select only the petal length and petal width columns, then arrange the data by petal width in ascending order.

```{r}
versicolor_sorted <- iris %>% dplyr::filter(variety == "versicolor") %>% dplyr::select(petal.length, petal.width) %>% dplyr::arrange(petal.width)
head(versicolor_sorted)
```
## Tidying data with tidyr

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyr)
```

#### **Wide Format**
In the wide format, each variable is spread out across multiple columns, and each observation (or unit) is represented by a single row.

**Example with the `iris` dataset:**
In its original wide format, the `iris` dataset looks like this:

| Species    | Sepal.Length | Sepal.Width | Petal.Length | Petal.Width |
|------------|--------------|-------------|--------------|-------------|
| setosa     | 5.1          | 3.5         | 1.4          | 0.2         |
| versicolor | 7.0          | 3.2         | 4.7          | 1.4         |
| virginica  | 6.3          | 3.3         | 6.0          | 2.5         |

Here, each column represents a different variable (e.g., `Sepal.Length`, `Sepal.Width`), and each row represents an observation.

#### **Long Format**
In the long format, each row corresponds to a single observation of a variable, with one column for the variable names and another for the variable values. This format is often more suitable for analysis and plotting, especially when using packages like `ggplot2`.

**Example of converting the `iris` dataset to long format:**

| Species    | Measurement   | Value |
|------------|---------------|-------|
| setosa     | Sepal.Length  | 5.1   |
| setosa     | Sepal.Width   | 3.5   |
| setosa     | Petal.Length  | 1.4   |
| setosa     | Petal.Width   | 0.2   |
| versicolor | Sepal.Length  | 7.0   |
| versicolor | Sepal.Width   | 3.2   |
| ...        | ...           | ...   |

Here, the variables `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and `Petal.Width` are consolidated into a single column called `Measurement`, with the corresponding values in another column called `Value`. Each unique combination of `Species` and `Measurement` has its own row.

#### **Differences and Use Cases**

- **Wide Format:**
  - Easier to view and understand when there are a small number of variables.
  - Each row represents a single unit or observation.
  - Suitable for certain types of statistical modeling or when data is naturally structured this way.

- **Long Format:**
  - More efficient for data manipulation, aggregation, and visualization, especially when using tools that work well with this structure (e.g., `ggplot2` for plotting).
  - Useful for datasets with multiple measurements per observation or when the number of variables is large.
  - Easier to filter, group, and apply functions across different variables.

### 1. Pivoting Longer
Convert the dataset into a longer format with columns measurement and value, where measurement includes the variable names (sepal length, sepal width, petal length, petal width) and value contains the corresponding measurements.

```{r}
# unify column names
iris <- iris %>% dplyr::rename(petal_length = petal.length, petal_width = petal.width, petal_ratio = petal.ratio)

# convert to long format
iris_long <- iris %>% tidyr::pivot_longer(cols = c(sepal_length, sepal_width, petal_length, petal_width),
               names_to = "Measurement",
               values_to = "Value")

head(iris_long)
```

### 2. Pivoting Wider
Take the dataset and pivot it back to original wide format.

```{r}
iris_wide <- iris_long %>% tidyr::pivot_wider(names_from = Measurement, values_from = Value)
head(iris_wide)
```

### 3. Separating Columns
Create a new dataset from iris where the variety column is split into two new columns: genus and species_type. Assume the first four letters represent the genus and the rest is the species type.

```{r}
iris_separated <- iris %>% tidyr::separate(variety, into = c("genus", "species_type"), sep = 4)
head(iris_separated)
```
### 4. Uniting Columns
Combine the sepal_length and sepal_width columns into a single column called sepal_dimensions with values separated by an underscore.

```{r}
iris_united <- iris %>% tidyr::unite(sepal_dimensions, sepal_length, sepal_width, sep = "_")
head(iris_united)
```

### 5. Dropping NA values
NA values are values missing from the dataset. Removal of NAs is a routine part of dataset pre-processing. In this exercise, we're gonna replace some values with NAs to obtain a more realistic dataset and then we're gonna remove rows with any missing values.

```{r}
set.seed(42) # for reproducibility

# introduce some NA values
iris_na <- iris %>% dplyr::mutate(sepal_length = replace(sepal_length, sample(1:n(), 10), NA))

# look at the introdued NAs
iris_na %>% dplyr::filter(is.na(sepal_length))
```
```{r}
iris_no_na <- iris_na %>% tidyr::drop_na()

# filter out rows with NAs in sepal_length
iris_no_na %>% dplyr::filter(is.na(sepal_length))
```

### 6. Replacing NA Values
In the iris_na dataset from previous example, replace the NA values in the sepal_length column with mean non-NA values of the sepal_length column.

```{r}
# calculate the man of the sepal_length column
mean_sepal_length <- mean(iris_na$sepal_length, na.rm = TRUE)
# replace NAs with the mean
iris_replaced_na <- iris_na %>% tidyr::replace_na(list(sepal_length = mean_sepal_length))
# check if NAs got replaced
iris_replaced_na %>% dplyr::filter(is.na(sepal_length))
```

## Data visualization with ggplot2

```{r, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
```

### 1. Basic Scatter Plot
Create a scatter plot of sepal length vs sepal width and color the points by variety.

```{r}
ggplot(iris, aes(x = sepal_length, y = sepal_width, color = variety)) + geom_point()
```

### 2. Adding Titles and Labels
Modify the scatter plot to include a title and x-axis and y-axis label.

```{r}
ggplot(iris, aes(x = sepal_length, y = sepal_width, color = variety)) +
  geom_point() +
  ggtitle("Sepal Length vs Sepal Width") +
  xlab("Sepal Length (cm)") +
  ylab("Sepal Width (cm)")
```

### 3. Histogram
Create a histogram of petal length and use different colors to represent different variety.

```{r}
ggplot(iris, aes(x = petal_length, fill = variety)) +
  geom_histogram(position = "dodge", bins = 30) +
  xlab("Petal length (cm)") +
  ylab("Count") +
  ggtitle("Histogram of Petal Length by Variety")
```

### 4. Box Plot
Create a box plot of sepal length for each variety.

```{r}
ggplot(iris, aes(x = variety, y = sepal_length, fill = variety)) +
  geom_boxplot() +
  ggtitle("Box Plot of Sepal Length by Variety") +
  xlab("Variety") +
  ylab("Sepal Length (cm)")
```

### 5. Facet Grid
Use facetting to create separate scatter plots of petal length vs. petal width for each variety.

```{r}
ggplot(iris, aes(x = petal_length, y = petal_width)) +
  geom_point() +
  facet_grid(. ~ variety) +
  ggtitle("Petal Dimensions by Variety") +
  xlab("Petal Length (cm)") +
  ylab("Petal Width (cm)")
```

### 6. Density Plot
Create a density plot for sepal width for each variety, using different colors.

```{r}
ggplot(iris, aes(x = sepal_width, fill = variety)) +
  geom_density(alpha = 0.5) + # the alpha parameter regulates color transparency and it ranges between 0.0 (fully transparent) and 1.0 (fully opaque)
  ggtitle("Density Plot of Sepal Width by Variety") +
  xlab("Sepal Width (cm)")
```

### 7. Line Plot
Generate a line plot to show the average petal length by variety.

```{r}
iris_avg <- iris %>%
  group_by(variety) %>%
  summarize(avg_petal_length = mean(petal_length))

ggplot(iris_avg, aes(x = variety, y = avg_petal_length, group = 1)) +
  geom_line() +
  geom_point() +
  ggtitle("Average Petal Length by Variety") +
  xlab("Variety") +
  ylab("Average Petal Length (cm)")
```

### 8. Bar Chart
Create a bar chart showing the count of observations for each variety.

```{r}
ggplot(iris, aes(x = variety, fill = variety)) +
  geom_bar() +
  ggtitle("Count of Observations by Variety") +
  xlab("Variety") +
  ylab("Count")
```

### 9. Customizing Plot Themes
Customize a scatter plot with sepal length vs petal length to use a minimal theme and change the legend position.

```{r}
ggplot(iris, aes(x = sepal_length, y = petal_length, color = variety)) +
  geom_point() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle("Sepal vs. Petal Length") +
  xlab("Sepal Length (cm)") +
  ylab("Petal Length (cm)")
```

## Working with categorical data with `forcats`

For this part we will use the `diamonds` dataset which has more columns with categorical data, namely cut, color, clarity.

```{r}
library(ggplot2)
diamonds$color
```

### 1. Reordering Factor Levels by Median Price

Reorder the `cut` factor levels in the `diamonds` dataset by the median `price` of diamonds in each `cut` category, in descending order.

```{r}
levels(diamonds$cut)

diamonds$cut <- forcats::fct_reorder(diamonds$cut, diamonds$price, .fun = median, .desc = TRUE)

levels(diamonds$cut)
```

### 2. Lump Rare Levels in `color`

Lump the least common levels of the `color` factor into an "Other" category, keeping only the top 4 most common colors distinct.

```{r}
levels(diamonds$color)

diamonds$color <- forcats::fct_lump_n(diamonds$color, n = 4)

levels(diamonds$color)
```

### 3. Collapsing Levels of `clarity`

Combine the `clarity` levels into broader categories:

- "Low" for I1, SI2, SI1
- "Medium" for VS2, VS1
- "High" for VVS2, VVS1, IF

```{r}
levels(diamonds$clarity)

diamonds$clarity <- forcats::fct_collapse(diamonds$clarity,
                                 Low = c("I1", "SI2", "SI1"),
                                 Medium = c("VS2", "VS1"),
                                 High = c("VVS2", "VVS1", "IF"))

levels(diamonds$clarity)
```

### 4. Renaming Factor Levels

Rename the cut factor levels to more descriptive names:

- "Fair" to "Fair Cut"
- "Good" to "Good Cut"
- "Very Good" to "Very Good Cut"
- "Premium" to "Premium Cut"
- "Ideal" to "Ideal Cut"

```{r}
diamonds$cut <- forcats::fct_recode(diamonds$cut,
                           "Fair Cut" = "Fair",
                           "Good Cut" = "Good",
                           "Very Good Cut" = "Very Good",
                           "Premium Cut" = "Premium",
                           "Ideal Cut" = "Ideal")

levels(diamonds$cut)
```

### 5. Reverse Order of Levels

Reverse the order of the levels in the `clarity` factor.

```{r}
levels(diamonds$clarity)

diamonds$clarity <- forcats::fct_rev(diamonds$clarity)

levels(diamonds$clarity)
```

### 6. Dropping Unused Levels

Create a subset of the diamonds `dataset` that only includes `Premium` and `Ideal` cuts, and then remove the unused levels from the `cut` factor.

```{r}
diamonds_subset <- diamonds %>%
  dplyr::filter(cut %in% c("Premium", "Ideal"))

diamonds_subset$cut <- forcats::fct_drop(diamonds_subset$cut)

levels(diamonds_subset$cut)
```

### 7. Convert a Numeric Variable to a Factor

Convert the `carat` variable into a categorical factor by binning it into intervals:

- 0.0-0.5 carats
- 0.5-1.0 carats
- 1.0-1.5 carats
- 1.5-2.0 carats
- 2.0+ carats

```{r}
diamonds$carat_bin <- cut(diamonds$carat, 
                          breaks = c(0, 0.5, 1, 1.5, 2, Inf), 
                          labels = c("0-0.5", "0.5-1.0", "1.0-1.5", "1.5-2.0", "2.0+"))

levels(diamonds$carat_bin)
```

## String Manipulation with `stringr`

```{r}
library(stringr)
```

### 1. Detecting Patterns in Strings

Find all rows in the `diamonds` dataset where the `cut` contains the substring "Good"

```{r}
good_cut_diamonds <- diamonds[stringr::str_detect(diamonds$cut, "Good"), ]
head(good_cut_diamonds)
```

### 2. Extracting Parts of a String

Create a string vector of diamond IDs in the format "Diamond_ID-XXXXX", where "XXXXX" is the row number. Then, extract just the numeric ID part from these strings.

```{r}
diamond_ids <- str_c("Diamond_ID-", rownames(diamonds))
extracted_ids <- stringr::str_extract(diamond_ids, "\\d+")
head(extracted_ids)
```

### 3. Replacing Substring

Replace all instances of "Good" in the `cut` column with "Decent"

```{r}
head(diamonds$cut)
diamonds$cut <- stringr::str_replace(diamonds$cut, "Good", "Decent")
head(diamonds$cut)
```

### 4. Splitting Strings

Split a hypothetical column of carat weights stored as a string with a unit (e.g., "1.5 carats") into numeric and unit components.

```{r}
carat_strings <- c("1.5 carats", "2.0 carats", "0.9 carats")
split_carat <- stringr::str_split(carat_strings, " ", simplify = TRUE)
carat_values <- as.numeric(split_carat[, 1])
carat_units <- split_carat[, 2]

head(carat_values)
head(carat_units)
```

### 5. String Length

Calculate the number of characters in each level of the `cut` column

```{r}
cut_lengths <- stringr::str_length(diamonds$cut)
head(cut_lengths)
```

### 6. Combining Strings

Combine the `cut`, `color` and `clarity` columns into a single string, separated by hyphensm to create a unique identifier for each diamond

```{r}
diamond_id <- stringr::str_c(diamonds$cut, diamonds$color, diamonds$clarity, sep = "-")
head(diamond_id)

```

### 7. Detecting Digits

Create a vector of mixed strings and identify which ones contain numeric digits

```{r}
mixed_strings <- c("carat 2.5", "weight", "price $500", "value 1000")
contains_digits <- stringr::str_detect(mixed_strings, "\\d")
contains_digits
```

### 8. Trimming Whitespace

Remove leading and trailing whitespace from a vector of strings representing diamond descriptions

```{r}
diamond_descriptions <- c("  Ideal Cut  ", "  Premium Quality ", "Fair Cut ")
trimmed_descriptions <- stringr::str_trim(diamond_descriptions)
trimmed_descriptions
```

### 9. Counting Substring Occurences

Count how many times the letter "e" appears in the `cut` column across the entire dataset

```{r}
e_count <- stringr::str_count(diamonds$cut, "e")
total_e_count <- sum(e_count)
total_e_count
```

### 10. Repeating Strings

Create a string that repeats the word "Diamond" five times, separated by spaces

```{r}
repeated_string <- stringr::str_dup("Diamond", 5)
repeated_string
```

### 11. Changing Case

Covert the `color` column in the `diamonds` dataset to uppercase

```{r}
head(diamonds$color)
diamonds$color <- stringr::str_to_upper(diamonds$color)
head(diamonds$color)
```

### 12. Subsetting Strings

Taken the first three characters from each value in the `clarity` column

```{r}
clarity_subset <- stringr::str_sub(diamonds$clarity, 1, 3)
head(clarity_subset)
```

### 13. Padding Strings

Create a column of diamond IDs (like in Exercise 2) and pad them with zeros on the left so that all IDs have five digits

```{r}
diamond_ids <- stringr::str_pad(row.names(diamonds), width = 5, pad = "0")
head(diamond_ids)
```

## Date and Time Manipualtion with `lubridate`

```{r}
library(lubridate)
```

### 1. Parsing Dates

Convert the string "2023-08-15" into a date object

```{r}
date <- lubridate::ymd("2023-08-15")
date
```

### 2. Extracting Date Components

From the date "2024-12-25", extract the year, month, and day

```{r}
date <- lubridate::ymd("2024-12-25")
year <- lubridate::year(date)
month <- lubridate::month(date)
day <- lubridate::day(date)

year
month
day
```

### 3. Handling Different Date Formats

Parse the date "15-August-2023" into a date object

```{r}
date <- lubridate::dmy("15-August-2023")
date
```

### 4. Adding and Subtracting Time Intervals

Add 3 months to the date "2023-01-01" and subtract 10 days from it

```{r}
date <- lubridate::ymd("2023-01-01")
date_plus_3_months <- date %m+% months(3)
date_minus_10_days <- date - days(10)

date_plus_3_months
date_minus_10_days
```

### 5. Calculating Time Differences

Calculate the number of days between "2023-01-01" and "2024-01-01"

```{r}
date1 <- lubridate::ymd("2023-01-01")
date2 <- lubridate::ymd("2024-01-01")
days_diff <- as.numeric(difftime(date2, date1, units = "days"))

days_diff
```

### 6. Working with Date-Time Objects

Create a date-time object for "2023-08-15 12:34:56"

```{r}
datetime <- lubridate::ymd_hms("2023-08-15 12:34:56")
datetime
```

### 7. Extracting Time Components

From the date-time "2023-08-15 12:34:56", extract the hour, minute, and second

```{r}
datetime <- lubridate::ymd_hms("2023-08-15 12:34:56")
hour <- lubridate::hour(datetime)
minute <- lubridate::minute(datetime)
second <- lubridate::second(datetime)

hour
minute
second
```

### 8. Rounding Dates

Round the date-time "2023-08-15 12:34:56" to the nearest hour

```{r}
datetime <- lubridate::ymd_hms("2023-08-15 12:34:56")
rounded_datetime <- lubridate::round_date(datetime, unit = "hour")

rounded_datetime
```

### 9. Finding the Day of the Week

Determine the day of the week for "2023-08-15"

```{r}
date <- lubridate::ymd("2023-08-15")
day_of_week <- lubridate::wday(date, label = TRUE)

day_of_week
```

### 10. Converting Time Zones

Convert the date-time "2023-08-15 12:34:56" from UTC to the "America/New_York" time zone

```{r}
datetime_utc <- lubridate::ymd_hms("2023-08-15 12:34:56", tz = "UTC")
datetime_ny <- lubridate::with_tz(datetime_utc, tzone = "America/New_York")

datetime_ny
```

## Modern reimagining of data frames with `tibble`

```{r}
library(tibble)
```

### 1. Creating a Simple Tibble

Create a tibble with three columns: `Name` (character), `Age` (numeric), and `Height` (numeric)

```{r}
people <- tibble::tibble(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Height = c(160, 175, 168)
)

people
```

### 2. Adding a New Column to a Tibble

Add a new column `Weight` to the `people` tibble created in Exercise 1

```{r}
people <- people %>%
  dplyr::mutate(Weight = c(55, 70, 68))

people
```

### 3. Converting a Data Frame to a Tibble

Convert the built-in `mtcars` data frame to a tibble.

```{r}
mtcars_tibble <- tibble::as_tibble(mtcars)
mtcars_tibble
```

### 4. Accessing Columns in a Tibble

Access the `mpg` column of the `mtcars_tibble` created in Exercise 3 using the `$` operator.

```{r}
mpg_column <- mtcars_tibble$mpg
mpg_column
```

### 5. Accessing Rows in a Tibble

Access the first 5 rows of the `mtcars_tibble`

```{r}
first_five_rows <- mtcars_tibble[1:5, ]
first_five_rows
```

### 6. Creating a Tibble with Different Data Types

Create a tibble with a date column, a logical column, and a character column

```{r}
mixed_tibble <- tibble::tibble(
  Date = as.Date(c("2023-01-01", "2023-01-02", "2023-01-03")),
  Logical = c(TRUE, FALSE, TRUE),
  Category = c("A", "B", "C")
)

mixed_tibble
```

### 7. Renaming Columns in a Tibble

Rename the columns in the `people` tibble to `Full_Name`, `Age_Years`, and `Height_cm`

```{r}
people <- people %>%
  dplyr::rename(
    Full_Name = Name,
    Age_Years = Age,
    Height_cm = Height
  )

people
```

### 8. Subsetting a Tibble with `filter` and `select`

Subset the `mtcars_tibble` to only include rows where `mpg` is greater than 20 and select only the `mpg`, `hp`, and `wt` columns

```{r}
filtered_tibble <- mtcars_tibble %>%
  dplyr::filter(mpg > 20) %>%
  dplyr::select(mpg, hp, wt)

filtered_tibble
```

## Functional Programming with `purrr`

```{r}
library(purrr)
```

### 1. Applying a Function to Each Element in a List

Create a list of numeric vectors and apply the `mean` function to each vector

```{r}
num_list <- list(c(1, 2, 3), c(4, 5, 6, 7), c(8, 9))
means <- purrr::map_dbl(num_list, mean)

means
```

### 2. Mapping Over Multiple Lists 

Given two lists of numeric vectors, sum corresponding elements of each vector

```{r}
list1 <- list(c(1, 2, 3), c(4, 5))
list2 <- list(c(10, 20, 30), c(40, 50))
sums <- purrr::map2(list1, list2, ~ .x + .x)

sums
```

### 3. Extracting Elements from a List of Lists

Extract the first element from each sublist in a list of lists

```{r}
nested_list <- list(
  list(a = 1, b = 2),
  list(a = 3, b = 4),
  list(a = 5, b = 6)
)
first_elements <- purrr::map(nested_list, "a")

first_elements
```

### 4. Safely Applying Functions to Avoid Errors

Use `safely` to apply a function that might fail (e.g., division) to a list of values, handling errors gracefully.

```{r}
safe_divide <- safely(~ 10 / .x)

results <- purrr::map(list(2, 0, 5), safe_divide)
results
```

### 5. Flattening a List of Lists

Flatten a list of lists into a single list

```{r}
nested_list <- list(list(1, 2), list(3, 4), list(5, 6))
flat_list <- purrr::flatten(nested_list)

flat_list
```

### 6. Creating a tibble from a List

Create a tibble from a list where each element is a vector of the same length

```{r}
list_data <- list(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Height = c(160, 175, 168)
)

people_tibble <- tibble::as_tibble(list_data)
people_tibble
```

### 7. Applying a Function Over Multiple Arguments

Given two numeric vectors, apply a function that calculates the maximum of each pair of elements

```{r}
vec1 <- c(1, 4, 3)
vec2 <- c(2, 3, 5)

max_values <- map2_dbl(vec1, vec2, max)
max_values
```

### 8. Mapping Functions to Columns of a Data Frame

Apply the `mean` function to each column in the `mtcars` dataset

```{r}
mtcars_means <- purrr::map_dbl(mtcars, mean)

mtcars_means
```

### 9. Applying a Function Conditionally

Apply a function to each element of a list only if the element meets a certain condition (e.g., is greater than 3)

```{r}
num_list <- list(1, 4, 3, 6, 2)

results <- purrr::map_if(num_list, ~ .x > 3, ~ .x * 2)
results
```

### 10. Reducing a List to a Single Value

Use `reduce` to calculate the cumulative product of a list of numbers

```{r}
num_list <- list(1, 2, 3, 4, 5)

product <- purrr::reduce(num_list, `*`)
product
```

## Functional Programming with `purrr`

```{r}
library(purrr)
```

### 1. Applying a Function to Each Element in a List

Create a list of numeric vectors and apply the `mean` function to each vector.

```{r}
# TODO your code here
```

### 2. Mapping Over Multiple Lists

Given two lists of numeric vectors, sum corresponding elements of each vector.

```{r}
# TODO your code here
```

### 3. Extracting Elements from a List of Lists

Extract the first element from each sublist in a list of lists.

```{r}
# TODO your code here
```

### 4. Safely Applying Functions to Avoid Errors

Use `safely` to apply a function that might fail (e.g., division) to a list of values, handling errors gracefully.

```{r}
# TODO your code here
```

### 5. Flattening a List of Lists

Flatten a list of lists into a single list.

```{r}
# TODO your code here
```

### 6. Creating a tibble from a List

Create a tibble from a list where each element is a vector of the same length.

```{r}
# TODO your code here
```

### 7. Applying a Function Over Multiple Arguments

Given two numeric vectors, apply a function that calculates the maximum of each pair of elements.

```{r}
# TODO your code here
```

### 8. Mapping Functions to Columns of a Data Frame

Apply the `mean` function to each column in the `mtcars` dataset.

```{r}
# TODO your code here
```

### 9. Applying a Function Conditionally

Apply a function to each element of a list only if the element meets a certain condition (e.g., is greater than 3).

```{r}
# TODO your code here
```

### 10. Reducing a List to a Single Value

Use `reduce` to calculate the cumulative product of a list of numbers.

```{r}
# TODO your code here
```