---
title: 'Tidyverse Data Manipulation: Exercises'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exercises

## Loading Data

### Exercise 1: Load the Data

**Task:** Use `read_csv()` to load the dataset from a file named `penguins.csv` and display the first 6 rows.

```{r}
library(readr)

penguins <- read_csv("penguins.csv")
head(penguins, 6)
```

## Install the complete `tidyverse` package

```{r}
install.packages("tidyverse")
library(tidyverse)
```

## Manipulating Data with dplyr

### Exercise 2: Filter Rows by Species

**Task:** Filter the dataset to include only rows where the species is "Adelie".

```{r}
penguins %>% filter(species == "Adelie")
```

### Exercise 3: Select Relevant Columns

**Task:** Select the columns `island`, `bill_length_mm`, and `flipper_length_mm` from the dataset.

```{r}
penguins %>% select(c("island", "bill_length_mm", "flipper_length_mm"))
```

### Exercise 4: Add a New Column

**Task:** Create a new column that calculates the ratio of `bill_length_mm` to `flipper_length_mm`.

```{r}
penguins %>% mutate(bill_to_flipper_ratio = bill_length_mm / flipper_length_mm)
```

### Exercise 5: Summarize Data by Species

**Task:** Calculate the mean and standard deviation of `flipper_length_mm` for each species.

```{r}
penguins %>% group_by(species) %>% summarize(mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
                                             sd_flipper = sd(flipper_length_mm, na.rm = TRUE))
```

### Exercise 6: Arrange Data by Bill Length

**Task:** Arrange the dataset by `bill_length_mm` in descending order.

```{r}
penguins %>% arrange(desc(bill_length_mm))
```

### Exercise 7: Group and Summarize by Species

**Task:** For each species, count how many instances have `bill_length_mm` greater than 40.

```{r}
penguins %>% group_by(species) %>% filter(bill_length_mm > 40) %>% count()
```

```{r}
penguins %>%
  group_by(species) %>%
  summarise(
    count_bill_gt40 = sum(bill_length_mm > 40, na.rm = TRUE)
  )
```

### Exercise 8: Rename Columns

**Task:** Rename the columns `bill_length_mm` and `flipper_length_mm` to `bill_length` and `flipper_length`.

```{r}
penguins %>% rename(bill_length = bill_length_mm, flipper_length = flipper_length_mm)
```

## Tidying Data with tidyr

### Exercise 9: Pivoting Longer

**Task:** Convert the dataset into a longer format with columns `measurement` and `value`, where `measurement` includes the variable names (`bill_length_mm`, `flipper_length_mm`) and `value` contains the corresponding measurements.

```{r}
library(tidyr)

penguins_long <- penguins %>%
  pivot_longer(
    cols = c(bill_length_mm, flipper_length_mm),
    names_to = "measurement",
    values_to = "value"
  )
```

### Exercise 10: Pivoting Wider

**Task:** Pivot the dataset back to the original wide format.

```{r}
penguins_wide <- penguins_long %>%
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

### Exercise 11: Separate a Column

**Task:** Split the `species` column into two new columns: `genus` and `species_type`.

```{r}
penguins %>%
  separate(species, into = c("genus", "species_type"), sep = " ")
```

### Exercise 12: Drop NA Values

**Task:** Remove rows with missing values from the dataset.

```{r}
penguins %>% drop_na()
```

## Working with Categorical Data with `forcats`

### Exercise 13: Lump Rare Species into "Other"

**Task:** Lump the least common species into an "Other" category, keeping only the top 3 most common species distinct.

```{r}
library(forcats)

penguins %>% mutate(species = fct_lump(species, n = 3))
```

### Exercise 14: Reorder Factor Levels

**Task:** Reorder the `species` factor by the mean `bill_length_mm` of each species in descending order.

```{r}
penguins %>% mutate(species = fct_reorder(species, bill_length_mm, .fun = mean, .desc = TRUE))
```

### Exercise 15: Reverse the Factor Levels Order

**Task:** Reverse the order of the levels in the `species` factor.

```{r}
penguins %>% mutate(species = fct_rev(species))
```

## String Manipulation with `stringr`

### Exercise 16: Detecting Substrings

**Task:** Find all rows in the `penguins` dataset where the `species` contains the substring "Adel".

```{r}
penguins %>% filter(str_detect(species, "Adel"))
```

### Exercise 17: Extract Numeric ID from String

**Task:** Create a string vector of penguin IDs in the format "Penguin_ID-XXXXX", where "XXXXX" is the row number. Extract just the numeric part of these IDs.

```{r}
penguin_ids <- paste0("Penguin_ID-", sprintf("%05d", seq_len(nrow(penguins))))
numeric_ids <- str_extract(penguin_ids, "\\d+")
```

## Date and Time Manipulation with `lubridate`

### Exercise 18: Parsing and Manipulating Dates

**Task:** Convert the string `"2023-08-15"` into a date object and extract the year, month, and day components.

```{r}
library(lubridate)

date <- ymd("2023-08-15")
year_val <- year(date)
month_val <- month(date)
day_val <- day(date)
```

### Exercise 19: Add and Subtract Time Intervals

**Task:** Add 3 months to the date `"2023-01-01"` and subtract 10 days from it.

```{r}
date <- ymd("2023-01-01")
date_plus_3_months <- date %m+% months(3)
date_minus_10_days <- date - days(10)
```

## Modern Reimagining of Data Frames with `tibble`

### Exercise 20: Create and Modify a Tibble

**Task:** Create a tibble with three columns: `Name` (character), `Age` (numeric), and `Height` (numeric). Then, add a new column `Weight`.

```{r}
library(tibble)

people <- tibble(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 22),
  Height = c(165, 180, 175)
)

people <- people %>%
  mutate(Weight = c(55, 72, 68))
```

### Exercise 21: Convert a Data Frame to Tibble

**Task:** Convert the built-in `mtcars` data frame to a tibble.

```{r}
as_tibble(mtcars)
```
