---
title: 'Central Limit Theorem: Exercises'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-09-26"
output: html_document
---

# Central Limit Theorem Exercises

## Exercise 1: Simulating a Population Distribution
Create a population distribution using a non-normal distribution (e.g., exponential or uniform). 
- Simulate a population of 10,000 values.
- Plot the population distribution.

```{r}
# TODO your code here
```

## Exercise 2: Sampling from the Population
Take random samples from the population you created in Exercise 1.
- Draw 30 random samples, each containing 50 observations.
- Calculate and store the means of these samples.

```{r}
# TODO your code here
```

## Exercise 3: Distribution of Sample Means
Plot the distribution of the sample means you calculated in Exercise 2.
- Use a histogram or density plot to visualize the distribution of sample means.

```{r}
# TODO your code here
```

## Exercise 4: Verifying the CLT
- Create a function to simulate sampling from the population and plotting the distribution of sample means.
- Test the function with different sample sizes (e.g., 10, 30, 100) and observe how the distribution of the sample means changes.

```{r}
# TODO your code here
```

## Exercise 5: Standard Error of the Mean
- Calculate the standard error of the mean (SEM) for your sample means from Exercise 2.
- Explain how SEM relates to the sample size and the variability of the population.

```{r}
# TODO your code here
```

## Exercise 6: Confidence Intervals
Using the sample means calculated in Exercise 2:
- Calculate a 95% confidence interval for the population mean based on your sample means.
- Interpret what this interval means in the context of your simulation.

```{r}
# TODO your code here
```

## Exercise 7: Real-World Application
Research a dataset (e.g., heights of individuals, test scores) and:
- Load the dataset into R and explore its distribution.
- Use the CLT to take samples and demonstrate the theorem by comparing the sample means distribution to the population distribution.

```{r}
# TODO your code here
```

## Exercise 8: Reflection
Write a brief reflection on how understanding the Central Limit Theorem is beneficial in real-world data analysis situations.

```{r}
# TODO your reflections here
```

