---
title: 'Central Limit Theorem: Exercises Solutions'
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-12-05"
output: html_document
---

# Central Limit Theorem Exercises Solutions

## Exercise 1: Simulating a Population Distribution
Create a population distribution using a non-normal distribution (e.g., exponential or uniform). 
- Simulate a population of 10,000 values.
- Plot the population distribution.

```{r}
library(ggplot2)
population_distribution <- rexp(10000, 1)
ggplot(as.data.frame(population_distribution), aes(x = population_distribution)) + geom_density(color = "#3498eb", fill = "#3498eb", alpha = 0.5)
```

## Exercise 2: Sampling from the Population
Take random samples from the population you created in Exercise 1.
- Draw 30 random samples, each containing 50 observations.
- Calculate and store the means of these samples.

```{r}
random_samples <- replicate(30, sample(population_distribution, size = 50, replace = TRUE))
random_samples_means <- apply(random_samples, 2, mean)
random_samples_means
```

## Exercise 3: Distribution of Sample Means
Plot the distribution of the sample means you calculated in Exercise 2.
- Use a histogram or density plot to visualize the distribution of sample means.

```{r}
ggplot(as.data.frame(random_samples_means), aes(x = random_samples_means)) + geom_density(color = "#3498eb", fill = "#3498eb", alpha = 0.5) + labs(title = "Distribution of Sample Means", x = "Sample Means", y = "Density")
```

## Exercise 4: Verifying the CLT
- Create a function to simulate sampling from the population and plotting the distribution of sample means.
- Test the function with different sample sizes (e.g., 10, 30, 100) and observe how the distribution of the sample means changes.

```{r}
plot_sample_means <- function(sample_size, num_samples, population_dist) {
  random_samples <- replicate(num_samples, sample(population_dist, size = sample_size, replace = TRUE))
  
  random_samples_means <- apply(random_samples, 2, mean)
  
  ggplot(as.data.frame(random_samples_means), aes(x = random_samples_means)) + geom_density(color = "#3498eb", fill = "#3498eb", alpha = 0.5) + labs(title = "Distribution of Sample Means", x = "Sample Means", y = "Density")
}

plot_sample_means(10, 1000, population_distribution)
plot_sample_means(30, 1000, population_distribution)
plot_sample_means(100, 1000, population_distribution)
```

## Exercise 5: Standard Error of the Mean
- Calculate the standard error of the mean (SEM) for your sample means from Exercise 2.
- Explain how SEM relates to the sample size and the variability of the population.

```{r}
random_samples_sds <- apply(random_samples, 2, sd)
n <- 50
# SEs <- random_samples_sds / sqrt(n)
SE <- sd(random_samples_means) / sqrt(n)
```

## Exercise 6: Confidence Intervals
Using the sample means calculated in Exercise 2:

  - Calculate a 95% confidence interval for the population mean based on your sample means.
  - Interpret what this interval means in the context of your simulation.

```{r}
# Calculate the mean and standard deviation of the sample means
mean_of_sample_means <- mean(random_samples_means)
sd_of_sample_means <- sd(random_samples_means)

# Number of samples
n_samples <- length(random_samples_means)

# Calculate the 95% confidence interval
z_score <- 1.96 # Z-score for 95% confidence level
margin_of_error <- z_score * (sd_of_sample_means / sqrt(n_samples))

lower_bound <- mean_of_sample_means - margin_of_error
upper_bound <- mean_of_sample_means + margin_of_error

# Output the confidence interval
confidence_interval <- c(lower_bound, upper_bound)
confidence_interval
```

## Exercise 7: Real-World Application
Research a dataset (e.g., heights of individuals, test scores) and:
- Load the dataset into R and explore its distribution.
- Use the CLT to take samples and demonstrate the theorem by comparing the sample means distribution to the population distribution.

```{r}
set.seed(42)
summary(iris$Sepal.Length)
ggplot(iris, aes(x=Sepal.Length)) + geom_density()

plot_sample_means(500, 50, iris$Sepal.Length)
```

## Exercise 8: Reflection
Write a brief reflection on how understanding the Central Limit Theorem is beneficial in real-world data analysis situations.

```{r}
# TODO your reflections here
```

