---
title: "t-Tests"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-10-17"
output: html_document
---

# t-Tests: One-Sample, Independent Two-Sample, and Paired t-Tests

## 1. **Introduction to t-Tests**
  - **What is a t-Test?**
    - A t-test is a statistical test used to determine whether there is a significant difference between the means of two groups or whether the mean of a single group is significantly different from a known value.
    - It is particularly useful when the sample size is small (\( n < 30 \)) and/or the population standard deviation is unknown.

  - **Types of t-Tests**:
    1. **One-Sample t-Test**: Compares the mean of a single sample to a known value (population mean).
    2. **Independent Two-Sample t-Test**: Compares the means of two independent groups.
    3. **Paired t-Test**: Compares means from the same group at different times (e.g., before and after treatment).

## 2. **Steps for a One-Sample t-Test**

1. **State the Hypotheses**:
    - **Null Hypothesis (\( H_0 \))**: The population mean is equal to a specific value.
    - Example: \( H_0 \): The mean height of a population is 170 cm.
    - **Alternative Hypothesis (\( H_a \))**: The population mean is not equal to that value (two-tailed), or it is greater or less than that value (one-tailed).

2. **Calculate the t-Statistic**:
    - Formula:
      \[
      t = \frac{\bar{x} - \mu_0}{\frac{s}{\sqrt{n}}}
      \]
      - \( \bar{x} \): sample mean
      - \( \mu_0 \): population mean (hypothesized)
      - \( s \): sample standard deviation
      - \( n \): sample size

3. **Find the Critical Value** (or p-value):
    - Use the `qt()` function to get the critical value based on the significance level \( \alpha \) and the degrees of freedom \( df = n - 1 \).
    - Alternatively, calculate the p-value using the `pt()` function.

4. **Make a Decision**:
    - Compare the t-statistic with the critical value or the p-value with \( \alpha \).
    - If \( |t| \) is greater than the critical value or p-value < \( \alpha \), **reject** the null hypothesis.
    - Otherwise, **fail to reject** the null hypothesis.

## 3. **Example: One-Sample t-Test in R**

Suppose we have a sample of 25 students with an average height of 172 cm. We want to test if the average height is different from 170 cm at a 5% significance level.

1. **Hypotheses**:
    - \( H_0 \): The mean height is 170 cm (\( \mu = 170 \)).
    - \( H_a \): The mean height is different from 170 cm (\( \mu \neq 170 \)).

2. **Calculate the t-statistic**:
```{r}
# Sample data
x_bar <- 172  # Sample mean
mu_0 <- 170   # Population mean under null hypothesis
s <- 8        # Sample standard deviation
n <- 25       # Sample size
alpha <- 0.05 # Significance level

# t-statistic
t <- (x_bar - mu_0) / (s / sqrt(n))
t
```

3. **Find the Critical Value and Make a Decision**:
```{r}
# Critical value for 95% confidence level (alpha = 0.05)
df <- n - 1  # Degrees of freedom
t_critical <- qt(1 - alpha/2, df)  # Two-tailed test
p_value <- 2 * (1 - pt(abs(t), df))
t_critical
p_value

# Decision: reject or fail to reject H0
if (abs(t) > t_critical) {
 print("Reject H0: There is a significant difference.")
} else {
 print("Fail to reject H0: There is no significant difference.")
}
```

## 4. **Independent Two-Sample t-Test**

An independent two-sample t-test is used to compare the means of two independent groups.

1. **Hypotheses**:
    - **Null Hypothesis (\( H_0 \))**: The means of the two groups are equal (\( \mu_1 = \mu_2 \)).
    - **Alternative Hypothesis (\( H_a \))**: The means of the two groups are not equal (\( \mu_1 \neq \mu_2 \)).

2. **Calculate the t-Statistic**:
   - Formula:
      \[
      t = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}
      \]
      where:
      - \( \bar{x}_1 \), \( \bar{x}_2 \): means of groups 1 and 2
      - \( s_1 \), \( s_2 \): standard deviations of groups 1 and 2
      - \( n_1 \), \( n_2 \): sample sizes of groups 1 and 2

3. **Example: Independent Two-Sample t-Test in R**:
```{r}
# Sample data
group1 <- c(10, 20, 30, 40, 50)  # Sample values for group 1
group2 <- c(15, 25, 35, 45, 55)  # Sample values for group 2

# t-test
t_test_result <- t.test(group1, group2, var.equal = TRUE)  # assuming equal variances
t_test_result
```

## 5. **Paired t-Test**

A paired t-test is used to compare means from the same group at different times (e.g., before and after treatment).

1. **Hypotheses**:
    - **Null Hypothesis (\( H_0 \))**: The mean difference between the paired observations is zero (\( \mu_d = 0 \)).
    - **Alternative Hypothesis (\( H_a \))**: The mean difference is not zero (\( \mu_d \neq 0 \)).

2. **Calculate the t-Statistic**:
   - Formula:
      \[
      t = \frac{\bar{d}}{\frac{s_d}{\sqrt{n}}}
      \]
      where:
      - \( \bar{d} \): mean of the differences
      - \( s_d \): standard deviation of the differences
      - \( n \): number of paired observations

3. **Example: Paired t-Test in R**:
```{r}
# Sample paired data
before <- c(80, 85, 88, 90, 92)  # Measurements before treatment
after <- c(82, 86, 89, 91, 94)   # Measurements after treatment

# Paired t-test
paired_test_result <- t.test(before, after, paired = TRUE)
paired_test_result
```

## 6. **Conclusion**
  - The **t-test** family of tests is a powerful tool for comparing means, with specific tests suited for different experimental designs (one-sample, two-sample, and paired).
  - By following the steps and examples, students will be able to conduct each type of t-test and interpret the results effectively.

## 7. **Term Vocabulary (Slovník pojmů)**

| **English Term**              | **Czech Translation**                        | **Explanation**                                                                            |
|-------------------------------|----------------------------------------------|--------------------------------------------------------------------------------------------|
| One-Sample t-Test             | Jednovýběrový t-test                        | A test that compares the mean of a single sample to a known population mean.               |
| Independent Two-Sample t-Test | Dvouvýběrový t-test                         | A test that compares the means of two independent groups.                                  |
| Paired t-Test                 | Párový t-test                               | A test that compares means from the same group at different times.                         |
| Null Hypothesis (\( H_0 \))    | Nulová hypotéza                             | The hypothesis that there is no effect or no difference.                                   |
| Alternative Hypothesis (\( H_a \)) | Alternativní hypotéza                  | The hypothesis that there is an effect or a difference.                                    |
| t-statistic                   | t-statistika                                | The test statistic used in a t-test.                                                       |
| Significance Level (\( \alpha \)) | Hladina významnosti (\( \alpha \))        | The probability of rejecting the null hypothesis when it is true (Type I error).            |
| p-value                       | p-hodnota                                   | The probability of obtaining the observed results, or more extreme ones, under \( H_0 \).   |
| Critical Value                | Kritická hodnota                            | The threshold value that the test statistic is compared to, based on the significance level