---
title: "t-Tests: Exercises"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-10-17"
output: html_document
---

# Exercises

## **Exercise 1: One-Sample t-Test**
1. A nutritionist claims that the average daily calorie intake of adults is 2,500 calories. A random sample of 30 adults shows an average intake of 2,450 calories with a standard deviation of 300 calories. Test the nutritionist's claim at the 5% significance level.
    - State the hypotheses.
    - Perform the one-sample t-test and make a decision.

```{r}
x_bar <- 2450  # Sample mean
mu_0 <- 2500   # Population mean under null hypothesis
s <- 300       # Sample standard deviation
n <- 30        # Sample size
alpha <- 0.05  # Significance level

# TODO Your code here
# t-statistic
# Critical value and p-value
# Decision
```

## **Exercise 2: Independent Two-Sample t-Test**
2. A researcher is investigating the effect of two different teaching methods on students' performance. Group A (30 students) has an average score of 85 with a standard deviation of 5, while Group B (35 students) has an average score of 80 with a standard deviation of 6. Test whether there is a significant difference between the two groups at a 5% significance level.

```{r}
x_bar_A <- 85   # Sample mean for Group A
s_A <- 5        # Sample standard deviation for Group A
n_A <- 30       # Sample size for Group A

x_bar_B <- 80   # Sample mean for Group B
s_B <- 6        # Sample standard deviation for Group B
n_B <- 35       # Sample size for Group B
alpha <- 0.05   # Significance level

# TODO Your code here
# t-test for two independent samples
# Critical value and p-value
# Decision
```

## **Exercise 3: Paired t-Test**
3. A psychologist wants to evaluate the effectiveness of a new therapy on anxiety levels. She measures the anxiety levels of 20 patients before and after the therapy. The differences in scores (after - before) are as follows: 5, 3, 8, 7, 2, 4, 6, 1, 9, 4, 2, 3, 5, 6, 7, 8, 4, 3, 5, 2. Test whether the therapy has a significant effect at a 1% significance level.

```{r}
differences <- c(5, 3, 8, 7, 2, 4, 6, 1, 9, 4, 2, 3, 5, 6, 7, 8, 4, 3, 5, 2)
n <- length(differences)  # Sample size
alpha <- 0.01  # Significance level

# TODO Your code here
# t-statistic for paired t-test
# Critical value and p-value
# Decision
```

## **Exercise 4: Interpretation of p-value**
4. In a study comparing the effects of two diets, a two-sample t-test yields a t-statistic of 2.45 and a p-value of 0.022. 
    - Based on a 5% significance level, would you reject or fail to reject the null hypothesis?
    - What would be your conclusion at a 1% significance level?

```{r}
t <- 2.45
p_value <- 0.022
alpha_5 <- 0.05
alpha_1 <- 0.01

# TODO Your interpretation here
# Decision at 5% significance level
# Decision at 1% significance level
```
