---
title: "Confidence Intervals: Exercises"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "2024-10-17"
output: html_document
---

# Exercises

## **Exercise 1: Confidence Interval for the Mean (Large Sample, Z-statistic)**

1. Given a sample of 50 randomly generated data points from a normal distribution with a mean of 65 and a standard deviation of 10, calculate the 95% confidence interval for the mean using the Z-statistic (assume n > 30).  
   
```{r}
set.seed(123)
sample_data <- rnorm(50, mean = 65, sd = 10)
# Calculate 95% CI using Z-statistic
# TODO Your code here
```

## **Exercise 2: Confidence Interval for the Mean (Small Sample, t-statistic)**

2. You have a small sample of 15 measurements from a population with an unknown mean and standard deviation. The sample values are:  
   `c(12, 15, 14, 11, 15, 18, 16, 14, 13, 19, 21, 16, 14, 12, 20)`.  
   Calculate the 95% confidence interval for the mean using the t-statistic (since n < 30).  
   
```{r}
sample_data <- c(12, 15, 14, 11, 15, 18, 16, 14, 13, 19, 21, 16, 14, 12, 20)
# Calculate 95% CI using t-statistic
# TODO Your code here
```

## **Exercise 3: Effect of Sample Size on Confidence Interval**

3. Generate two samples, one of size 20 and another of size 100, from a normal distribution with a mean of 50 and a standard deviation of 8. Calculate the 95% confidence interval for the mean of both samples.  
   Compare the widths of the two intervals and explain how sample size affects the confidence interval.

```{r}
set.seed(456)
sample_20 <- rnorm(20, mean = 50, sd = 8)
sample_100 <- rnorm(100, mean = 50, sd = 8)
# Calculate 95% CI for both sample_20 and sample_100
# TODO Your code here
```

## **Exercise 4: Visualizing Confidence Intervals with ggplot2**

4. Using a sample of 100 data points generated from a normal distribution (mean = 70, sd = 12), generate 1000 bootstrapped samples of the mean and create a histogram. Plot the 95% confidence interval for the bootstrapped means on the histogram using `ggplot2`.

```{r}
library(ggplot2)
set.seed(789)
sample_data <- rnorm(100, mean = 70, sd = 12)
# Bootstrap and plot 95% CI for the means
# TODO Your code here
```

## **Exercise 5: Interpretation of Confidence Intervals**

5. You calculated a 95% confidence interval for the mean of a dataset and obtained the interval [40.5, 49.5].  
   a. Interpret this result in the context of confidence intervals.  
   b. If the confidence level is increased to 99%, what do you expect will happen to the width of the interval?

```{r}
# TODO Your interpretation here
```

## **Exercise 6: Calculating Confidence Intervals with Built-in Functions**

6. Use R's built-in `t.test()` function to calculate the confidence interval for the following dataset:  
   `c(5.1, 5.4, 5.9, 6.2, 5.8, 5.7, 6.1, 5.6, 5.5, 6.3)`.  
   Calculate the 99% confidence interval for the mean of the data. 

```{r}
data <- c(5.1, 5.4, 5.9, 6.2, 5.8, 5.7, 6.1, 5.6, 5.5, 6.3)
# Calculate 99% CI using t.test()
# TODO Your code here
```
