---
title: "Confidence Intervals: Exercises Solutions"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

# Exercises

## **Exercise 1: Confidence Interval for the Mean (Large Sample, Z-statistic)**

1. Given a sample of 50 randomly generated data points from a normal distribution with a mean of 65 and a standard deviation of 10, calculate the 95% confidence interval for the mean using the Z-statistic (assume n > 30).  
   
```{r}
set.seed(123)
sample_data <- rnorm(50, mean = 65, sd = 10)

# Calculate 95% CI using Z-statistic

sample_data_mean <- mean(sample_data)
sample_data_sd <- sd(sample_data)

print(sample_data_mean)
sample_data_mean - qnorm(0.975) * sample_data_sd/sqrt(50)
sample_data_mean + qnorm(0.975) * sample_data_sd/sqrt(50)
```

## **Exercise 2: Confidence Interval for the Mean (Small Sample, t-statistic)**

2. You have a small sample of 15 measurements from a population with an unknown mean and standard deviation. The sample values are:  
   `c(12, 15, 14, 11, 15, 18, 16, 14, 13, 19, 21, 16, 14, 12, 20)`.  
   Calculate the 95% confidence interval for the mean using the t-statistic (since n < 30).  
   
```{r}
sample_data <- c(12, 15, 14, 11, 15, 18, 16, 14, 13, 19, 21, 16, 14, 12, 20)

# Calculate 95% CI using t-statistic

t.test(sample_data)$conf.int
```

## **Exercise 3: Effect of Sample Size on Confidence Interval**

3. Generate two samples, one of size 20 and another of size 100, from a normal distribution with a mean of 50 and a standard deviation of 8. Calculate the 95% confidence interval for the mean of both samples.  
   Compare the widths of the two intervals and explain how sample size affects the confidence interval.

```{r}
set.seed(456)
sample_20 <- rnorm(20, mean = 50, sd = 8)
sample_100 <- rnorm(100, mean = 50, sd = 8)

# Calculate 95% CI for both sample_20 and sample_100

t.test(sample_20)$conf.int

sample_100_mean <- mean(sample_100)
sample_100_sd <- sd(sample_100)

print(sample_100_mean - qnorm(0.975) * (sample_100_sd / sqrt(100)))
print(sample_100_mean + qnorm(0.975) * (sample_100_sd / sqrt(100)))
```

## **Exercise 4: Visualizing Confidence Intervals with ggplot2**

4. Using a sample of 100 data points generated from a normal distribution (mean = 70, sd = 12), generate 1000 bootstrapped samples of the mean and create a histogram. Plot the 95% confidence interval for the bootstrapped means on the histogram using `ggplot2`.

```{r}
library(ggplot2)
set.seed(789)
sample_data <- rnorm(100, mean = 70, sd = 12)

# Bootstrap and plot 95% CI for the means

boot_means <- replicate(1000, mean(sample(sample_data, size = 100, replace = TRUE)))
ci <- quantile(boot_means, c(0.025, 0.975))

# Visualizing CI
ggplot(data.frame(x = boot_means), aes(x = x)) +
 geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
 geom_vline(xintercept = ci, color = "red", linetype = "dashed") +
 labs(title = "Bootstrap Distribution with 95% Confidence Interval", x = "Sample Means")
```

## **Exercise 5: Interpretation of Confidence Intervals**

5. You calculated a 95% confidence interval for the mean of a dataset and obtained the interval [40.5, 49.5].  
   a. Interpret this result in the context of confidence intervals.  
   b. If the confidence level is increased to 99%, what do you expect will happen to the width of the interval?

```{r}
# a. We can be 95% confident, that the population mean lies somewhere in the range [40.5, 49.5].
# b. The width would increase. If we want higher confidence, then we need to cover wider range.
```

## **Exercise 6: Calculating Confidence Intervals with Built-in Functions**

6. Use R's built-in `t.test()` function to calculate the confidence interval for the following dataset:  
   `c(5.1, 5.4, 5.9, 6.2, 5.8, 5.7, 6.1, 5.6, 5.5, 6.3)`.  
   Calculate the 99% confidence interval for the mean of the data. 

```{r}
data <- c(5.1, 5.4, 5.9, 6.2, 5.8, 5.7, 6.1, 5.6, 5.5, 6.3)

# Calculate 99% CI using t.test()

t.test(data, conf.level=0.99)$conf.int
```
