---
title: "Random Variable: Exercises Solved"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

### **Exercise 1: Simulating a Biased Coin Toss**
Simulate 1000 tosses of a biased coin where the probability of getting heads is 0.7. Calculate and plot the Probability Mass Function (PMF) for the number of heads.

```{r}
set.seed(42)
library(ggplot2)

# simulate coin tosses
coin_tosses <- sample(c("heads", "tails"), size = 1000, replace = TRUE, prob = c(0.7, 0.3))

# plot PMF
coin_tosses_table <- table(coin_tosses)
coin_tosses_data <- as.data.frame(coin_tosses_table)
colnames(coin_tosses_data) <- c("coin_toss", "frequency")
coin_tosses_data$coin_toss <- as.factor(coin_tosses_data$coin_toss)
coin_tosses_data$probability <- coin_tosses_data$frequency / sum(coin_tosses_data$frequency)

ggplot(coin_tosses_data, aes(x = coin_toss, y = probability)) +
  geom_bar(stat = "identity", fill = "darkorange", color = "black") +
  labs(title = "Probability Mass Function of Coin Tosses",
       x = "Coin Toss Outcome",
       y = "Probability") +
  theme_minimal()
```

### **Exercise 2: CDF of a Biased Coin Toss**
Using the result from Exercise 1, calculate and plot the Cumulative Distribution Function (CDF) of the number of heads in 1000 tosses of the biased coin.

```{r}
set.seed(42)
# compute cumulative probabilities
cdf <- cumsum(coin_tosses_table) / sum(coin_tosses_table)

# Add a zero-probability point
cdf <- c(0, cdf)
outcomes <- c("Start", names(coin_tosses_table))  # add a label for the zero point

# create a data frame for plotting
cdf_df <- data.frame(
  outcome = names(cdf),
  cumulative_probability = as.numeric(cdf)
)

# convert outcome to factor and use numeric positions for x
cdf_df$numeric_outcome <- seq_along(cdf_df$outcome)

# plot using ggplot2
ggplot(cdf_df, aes(x = numeric_outcome, y = cumulative_probability)) +
  # horizontal lines for steps
  geom_segment(
    aes(x = numeric_outcome - 0.5, xend = numeric_outcome + 0.5,
        y = cumulative_probability, yend = cumulative_probability),
    color = "black"
  ) +
  # filled points at the left end
  geom_point(
    aes(x = numeric_outcome - 0.5, y = cumulative_probability),
    size = 3, color = "black", shape = 16  # filled circle
  ) +
  # hollow points at the right end
  geom_point(
    aes(x = numeric_outcome + 0.5, y = cumulative_probability),
    size = 3, color = "black", shape = 1  # hollow circle
  ) +
  # customize x-axis labels
  scale_x_continuous(breaks = cdf_df$numeric_outcome, labels = cdf_df$outcome) +
  labs(
    title = "Discontinuous CDF of Coin Toss Outcomes",
    x = "Outcome",
    y = "Cumulative Probability"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1))  # ensure y-axis spans from 0 to 1
```

### **Exercise 3: Custom Discrete Random Variable - Number of Goals in a Soccer Match**
Suppose a soccer team scores between 0 and 4 goals per match, with the following probabilities:  
- 0 goals: 0.2  
- 1 goal: 0.3  
- 2 goals: 0.25  
- 3 goals: 0.15  
- 4 goals: 0.1  

Write code to represent this custom discrete random variable. Then, calculate and plot its PMF and CDF.

```{r}
set.seed(42)
scores <- data.frame(number_of_goals = c(0:4), probability = c(0.2, 0.3, 0.25, 0.15, 0.1))
# plot PMF
ggplot(scores, aes(x = number_of_goals, y = probability)) +
  geom_bar(stat = "identity", fill = "yellowgreen", color = "black") +
  labs(title = "Probability Mass Function of Soccer Match Scores",
       x = "Number of Goals in a Match",
       y = "Probability") +
  theme_minimal()

# compute the cumulative probabilities (CDF)
scores$cumulative_probability <- cumsum(scores$probability)

# plot the CDF using ggplot2
ggplot(scores, aes(x = number_of_goals, y = cumulative_probability)) +
  # horizontal lines for steps
  geom_segment(
    aes(x = number_of_goals - 0.5, xend = number_of_goals + 0.5,
        y = cumulative_probability, yend = cumulative_probability),
    color = "black"
  ) +
  # filled points at the left end
  geom_point(
    aes(x = number_of_goals - 0.5, y = cumulative_probability),
    size = 3, color = "black", shape = 16  # filled circle
  ) +
  # hollow points at the right end
  geom_point(
    aes(x = number_of_goals + 0.5, y = cumulative_probability),
    size = 3, color = "black", shape = 1  # hollow circle
  ) +
  labs(
    title = "Discontinuous CDF of Goals in a Soccer Match",
    x = "Outcome",
    y = "Cumulative Probability"
  ) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1))  # ensure y-axis spans from 0 to 1
```

### **Exercise 4: Simulating Normally Distributed Test Scores**
Simulate the scores of 500 students on an exam. The test scores are normally distributed with a mean of 75 and a standard deviation of 10.  
1. Calculate the probability of a student scoring above 90.  
2. Plot the Probability Density Function (PDF) of the scores.

```{r}
set.seed(42)
mean_score <- 75
sd_score <- 10
test_scores <- rnorm(500, mean = mean_score, sd = sd_score)

# calculate the probability of scoring above 90
prob_above_90 <- 1 - pnorm(90, mean = mean_score, sd = sd_score)
prob_above_90

# plot the PDF of the test scores
ggplot(data.frame(scores = test_scores), aes(x = scores)) +
  geom_density(fill = "yellowgreen", alpha = 0.5) +
  geom_vline(xintercept = 90, linetype = "dashed", color = "red") +
  labs(title = "Probability Distribution of Simulated Exam Scores",
       x = "Score",
       y = "Density") +
  theme_minimal()
```

### **Exercise 5: Estimating CDF from Empirical Data**
Using the test scores simulated in Exercise 4, calculate and plot the empirical Cumulative Distribution Function (CDF). Estimate the probability that a student scores less than 60.

```{r}
set.seed(42)
# calculate the empirical CDF and the probability of scoring less than 60
test_scores_ecdf <- ecdf(test_scores)
prob_less_than_60 <- test_scores_ecdf(60)
prob_less_than_60

# plot the empirical CDF
ggplot(data.frame(x = test_scores), aes(x)) +
  stat_ecdf(geom = "step", color = "blue", size = 1.2) +
  geom_vline(xintercept = 60, linetype = "dashed", color = "red") +
  annotate("text", x = 60, y = 0.1, label = paste("P(Score < 60) =", round(prob_less_than_60, 4)), color = "red", hjust = -0.1) +
  labs(title = "Empirical Cumulative Distribution Function (CDF)",
       x = "Score",
       y = "Cumulative Probability") +
  theme_minimal()
```

### **Exercise 6: Exploring the Relationship Between PMF and Expected Value**
A company sells a product for $5. Each sale has a probability of 0.1, and they contact 100 customers.  
1. Simulate the number of sales made out of 100 customers (binomial distribution).
2. Calculate the expected revenue from these sales.
3. Plot the PMF of the number of sales.

```{r}
set.seed(42)
# parameters
n_customers <- 100    # number of customers contacted
p_sale <- 0.1         # probability of a sale

# simulate number of sales
sales <- rbinom(1, size = 100, prob = 0.1)
sales

# calculate the expected revenue
revenue <- sales * 5
revenue

# plot the PMF of the number of sales
pmf_data <- data.frame(sales = 0:100, 
                       probability = dbinom(0:100, size = 100, prob = 0.1))

ggplot(pmf_data, aes(x = sales, y = probability)) +
  geom_bar(stat = "identity", fill = "hotpink1", color = "black") +
  labs(title = "PMF of Number of Sales",
       x = "Number of Sales",
       y = "Probability") +
  theme_minimal()
```

### **Exercise 7: Comparing Two Continuous Random Variables**
Simulate two continuous random variables:  
1. X ~ N(0, 1)  
2. Y ~ N(3, 1.5)  

1. Plot the Probability Density Functions (PDFs) of both variables on the same graph.
2. Calculate the probability that X is greater than Y.

```{r}
set.seed(42)
library(dplyr)
library(tidyr)

# X ~ N(0, 1) and Y ~ N(3, 1.5)
n <- 1000
X <- rnorm(n, mean = 0, sd = 1)
Y <- rnorm(n, mean = 3, sd = 1.5)

# plot the PDFs of X and Y on the same plot
pdf_data <- data.frame(
  value = c(X, Y),
  variable = c(rep("X ~ N(0,1)", n), rep("Y ~ N(3,1.5)", n))
)

ggplot(pdf_data, aes(x = value, fill = variable)) +
  geom_density(alpha = 0.5) +
  labs(title = "PDFs of X and Y",
       x = "Value",
       y = "Density") +
  theme_minimal()

# calculate the probability that X > Y
prob_X_greater_Y <- mean(X > Y)
prob_X_greater_Y
```


### **Exercise 8: Uniform Distribution - Random Walk**
Generate 1000 steps of a random walk, where each step is a draw from a uniform distribution between -1 and 1.  
1. Plot the trajectory of the random walk.
2. Calculate the cumulative sum of the walk and plot it.

```{r}
set.seed(42)
n_steps <- 1000
steps <- runif(n_steps, min = -1, max = 1)

# calculate the cumulative sum (random walk)
random_walk <- cumsum(steps)

# plot the trajectory of the random walk
step_data <- data.frame(step = 1:n_steps, position = steps)
ggplot(step_data, aes(x = step, y = position)) +
  geom_point(color = "deeppink4") +
  labs(title = "Trajectory of Random Walk Steps",
       x = "Step",
       y = "Step Value") +
  theme_minimal()

# cumulative sum plot
walk_data <- data.frame(step = 1:n_steps, position = random_walk)
ggplot(walk_data, aes(x = step, y = position)) +
  geom_line(color = "dodgerblue") +
  labs(title = "Cumulative Sum of Random Walk",
       x = "Step",
       y = "Cumulative Position") +
  theme_minimal()
```
