---
title: "Random Variable: Exercises Solved"
author: "Kateřina Večerková vecerkok@vscht.cz"
date: "`r Sys.Date()`"
output: html_document
---

### **Exercise 1: Simulating a Biased Coin Toss**
Simulate 1000 tosses of a biased coin where the probability of getting heads is 0.7. Calculate and plot the Probability Mass Function (PMF) for the number of heads.

```{r}
set.seed(42)
library(ggplot2)
# plot PMF
```

### **Exercise 2: CDF of a Biased Coin Toss**
Using the result from Exercise 1, calculate and plot the Cumulative Distribution Function (CDF) of the number of heads in 1000 tosses of the biased coin.

```{r}
# plot the CDF using a step plot
```

### **Exercise 3: Custom Discrete Random Variable - Number of Goals in a Soccer Match**
Suppose a soccer team scores between 0 and 4 goals per match, with the following probabilities:  
- 0 goals: 0.2  
- 1 goal: 0.3  
- 2 goals: 0.25  
- 3 goals: 0.15  
- 4 goals: 0.1  

Write code to represent this custom discrete random variable. Then, calculate and plot its PMF and CDF.

```{r}
# plot PMF
# plot CDF
```

### **Exercise 4: Simulating Normally Distributed Test Scores**
Simulate the scores of 500 students on an exam. The test scores are normally distributed with a mean of 75 and a standard deviation of 10.  
1. Calculate the probability of a student scoring above 90.  
2. Plot the Probability Density Function (PDF) of the scores.

```{r}
test_scores <- rnorm(500, mean = 75, sd = 10)
# Calculate the probability of scoring above 90
# Plot the PDF of the test scores
```

### **Exercise 5: Estimating CDF from Empirical Data**
Using the test scores simulated in Exercise 4, calculate and plot the empirical Cumulative Distribution Function (CDF). Estimate the probability that a student scores less than 60.

```{r}
# TODO your code here
```

### **Exercise 6: Exploring the Relationship Between PMF and Expected Value**
A company sells a product for $5. Each sale has a probability of 0.1, and they contact 100 customers.  
1. Simulate the number of sales made out of 100 customers.
2. Calculate the expected revenue from these sales.
3. Plot the PMF of the number of sales.

```{r}
# parameters
n_customers <- 100    # Number of customers contacted
p_sale <- 0.1         # Probability of a sale

# sales range (0 to 100 possible sales)
# compute the PMF for each possible number of sales
# display the PMF values
# check that the probabilities sum to 1 (as they should)
# calculate the expected revenue (revenue = 5 * number of sales)
# plot the PMF for the number of sales
# plot the PMF
# find the number of sales with the highest probability (most likely outcome)
```

### **Exercise 7: Comparing Two Continuous Random Variables**
Simulate two continuous random variables:  
1. X ~ N(0, 1)  
2. Y ~ N(3, 1.5)  

1. Plot the Probability Density Functions (PDFs) of both variables on the same graph.
2. Calculate the probability that X is greater than Y.

```{r}
library(dplyr)
library(tidyr)

# X ~ N(0, 1) and Y ~ N(3, 1.5)
# plot the PDFs of X and Y on the same plot
# calculate the probability that X > Y
# we need to find the distribution of the difference Z = X - Y and calculate P(Z > 0)
# P(Z > 0)
```


### **Exercise 8: Uniform Distribution - Random Walk**
Generate 1000 steps of a random walk, where each step is a draw from a uniform distribution between -1 and 1.  
1. Plot the trajectory of the random walk.
2. Calculate the cumulative sum of the walk and plot it.

```{r}
n_steps <- 1000
steps <- runif(n_steps, min = -1, max = 1)
# calculate the cumulative sum for the random walk trajectory
```
